---
title: "Finalizing plots"
description: "tips for making your plots readable and professional"
categories: [data visualization]
format:
  html: 
    code-fold: true
---

# Why do we care about how plots, figures, graphs, etc. look?

Data visualization is a _huge_ part of data storytelling, one of the core parts of being a data scientist. This is _especially_ relevant to environmental science: you're responsible for communicating not just about the environment, but what evidence (i.e. data) supports your claim. Therefore, it is crucial that environmental scientists communicate about their data clearly and effectively.  

In this class, your plots will be assessed using three (very broad) criteria:  
1. **accuracy**: is your plot accurately and _truthfully_ representing the data?  
2. **clarity**: is your plot clearly representing a pattern, relationship, message?  
3. **aesthetics**: does your plot look good?  

# General rules for good-looking plots

_(adapted from Allison Horst)_

## Non-negotiable (if you are missing these, you will not get full credit for your plot)

- Axes must have **complete** labels with units (very few exceptions to this)
    - for example: body_mass_g should be "Body mass (g)"
- Underlying data must always be shown if there is any summary (e.g. mean or median) or model prediction shown
- Underlying data should be displayed in such a way that points can be distinguished (you can do this with transparency and/or modifying the shape type)
- Colors must be different from the `ggplot` defaults
- theme must be different from the `ggplot` defaults
- your plot size should allow for all components of your plot (title, axis text, legend, etc.) to be visible and readable

## Additional points  

- logical start and end values of x or y axes (these are usually by default in `ggplot`, but you should double check)
- text labels should be large enough to see/read clearly
- be aware of color-blind friendly color palettes
- use one font throughout a plot

In general, **the simpler you can make a plot, the better.**

# Bad/better examples

Some examples of bad and better plots follow using data from the `{lterdatasampler}` package.

```{r}
#| message: false
#| echo: false

library(lterdatasampler)
library(tidyverse)
```

## Bar chart

```{r fig.width = 8, fig.height = 6}
and_vertebrates |> 
  # filter for cutthroat trout and cascade/pool/side channel/rapid location
  filter(species == "Cutthroat trout" & unittype %in% c("C", "P", "SC", "R")) |> 
  # group by location
  group_by(unittype) |> 
  # count number of observations
  count() |> 
  # plot bar chart of number of observations by unittype
  ggplot(mapping = aes(x = unittype,
                       y = n)) +
  geom_col() 
```

**Why is this bad?**  

- gap between bottom of bars and x-axis  
- meaningless y axis  
- gray background against gray bars and black text is hard to see  
- gridlines don't do much

```{r fig.width = 8, fig.height = 6, warning = FALSE}
and_vertebrates |> 
  # same filtering and summarizing steps as above
  filter(species == "Cutthroat trout" & unittype %in% c("C", "P", "SC", "R")) |> 
  group_by(unittype) |> 
  count() |> 
  ungroup() |> 
  # mutating unittype column to have full names of locations
  mutate(unittype = recode_values(
    unittype,
    "C" ~ "Cascade",
    "P" ~ "Pool",
    "SC" ~ "Side channel",
    "R" ~ "Rapid"
  )) |> 
  # plot bar chart of number of observations by unittype
  # reorder() in aes() reorders x-axis in order of descending n (count)
  ggplot(mapping = aes(x = reorder(unittype, -n),
                       y = n)) +
  geom_col(fill = "lightblue4", color = "#000000") +
  # expand takes away the gap at the bottom and at the top of the plot
  # limits sets the limits of the axis
  scale_y_continuous(expand = c(0, 0), limits = c(0, 12000)) +
  # change titles to be meaningful
  labs(x = "Sampling location", 
       y = "Count",
       title = "Number of Cutthroat trout observations differs across sampling locations") +
  # one of the built-in themes in ggplot
  theme_bw() +
  theme(# changing text sizes
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 14),
        strip.text = element_text(size = 14),
        # getting rid of gridlines
        panel.grid = element_blank(),
        # making the plot title bigger and centering it
        plot.title = element_text(size = 20, hjust = 0.5),
        plot.title.position = "plot",
        text = element_text(family = "Garamond")
        ) 
```

**Why is this better?**  

- text is bigger  
- gridlines are gone (because the theme is different from the default) 
- easier to see columns against background  
- complete axes  
- x-axis labels are complete ("Cascade" instead of "C")
- fill is different from default (demonstrates some level of customization)
- x-axis is reordered so that the differences across groups are easier to see


## Scatterplot

```{r fig.width = 8, fig.height = 6, warning = FALSE}

knz_bison |> 
  # calculate age of individual
  mutate(age = rec_year - animal_yob) |> 
  # plotting weight as a function of age 
  ggplot(mapping = aes(x = age,
                       y = animal_weight)) +
  # scatterplot
  geom_point()

```

**Why is this bad?**  

- grey background, black dots  
- hides some meaningful variation across sexes (male bison tend to be larger than female bison, but female bison tend to live longer)    
- axes are meaningless  
- small text size  
- points likely overlap, so some parts of the data are hidden

```{r fig.width = 8, fig.height = 6, warning = FALSE}
knz_bison |> 
  # calculate age of individual
  mutate(age = rec_year - animal_yob) |> 
  # plotting weight as a function of age 
  # coloring and shaping points by sex (male or female)
  ggplot(mapping = aes(x = age,
                       y = animal_weight,
                       color = animal_sex, 
                       shape = animal_sex)) +
  # scatterplot
  # points are transparent and large
  geom_point(alpha = 0.4,
             size = 2) +
  # manually setting open shapes and colors
  scale_shape_manual(values = c(21, 24)) +
  scale_color_manual(values = c("darkgreen", "cornflowerblue")) +
  # relabelling axes and legend title
  labs(x = "Bison age",
       y = "Recorded weight (lbs)",
       color = "Sex",
       shape = "Sex") +
  # another ggplot built-in theme
  theme_classic() +
  theme(# putting legend in plot area
        legend.position = c(0.85, 0.85),
        # legend text sizes
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        # text size, position, and font adjustment
        axis.text = element_text(size = 14), 
        axis.title = element_text(size = 16),
        plot.title = element_text(size = 18, hjust = 0.5),
        plot.title.position = "plot",
        text = element_text(family = "Garamond")
    )
```

**Why is this better?**  

- white background, no grid lines  
- points are shaped and colored by sex (and larger overall), so you can easily see the differences between groups    
- transparency shows overlapping points  
- complete axis labels    
- text is larger and font is changed  
- legend is in plot area (if there's space to do this, generally good)

## Boxplots

```{r fig.width = 8, fig.height = 6, warning = FALSE}
arc_weather |> 
  # extracting month from date column
  mutate(month = month(date, label = TRUE)) |> 
  # filtering to only include dates from 2018
  filter(date > as.Date("2017-12-31")) |> 
  # plotting mean daily air temperature
  ggplot(mapping = aes(x = month,
                       y = mean_airtemp)) +
  # boxplot (shows median mean daily air temperature)
  geom_boxplot()
```

**Why is this bad?**  

- grey background, black dots  
- doesn't show the underlying data 
- y-axis is meaningless  
- small text size  

```{r}
#| fig-width: 8
#| fig-height: 6

arc_weather |> 
  # extracting month from date column
  mutate(month = month(date, label = TRUE)) |> 
  # filtering to only include dates from 2018
  filter(date > as.Date("2017-12-31")) |> 
  # plotting mean daily air temperature
  # coloring and filling by month
  ggplot(mapping = aes(x = month,
                       y = mean_airtemp,
                       fill = month,
                       color = month)) +
  # make the boxplots narrower
  # take out outliers (redundant if you are showing underlying data)
  # making the outlines black so that they are easier to see
  geom_boxplot(width = 0.2,
               outliers = FALSE,
               color = "black") +
  # showing the underlying data
  # making sure the points aren't jittered along the y-axis
  geom_jitter(alpha = 0.5,
              height = 0,
              width = 0.4) +
  # specify the colors you want to use for each month
  scale_fill_manual(values = NatParksPalettes::natparks.pals(name = "Glacier",
                                                             n = 12,
                                                             type = "continuous")) + 
  scale_color_manual(values = NatParksPalettes::natparks.pals(name = "Glacier",
                                                              n = 12,
                                                              type = "continuous")) +
  # relabel the axis titles, plot title, and caption
  labs(x = "Month", 
       y = "Air temperature (\U00B0 C)",
       title = "Monthly temperatures at Toolik Field Station, Alaska") +
  # themes built in to ggplot
  theme_bw() +
  # other theme adjustments
  theme(legend.position = "none", 
        axis.title = element_text(size = 13),
        axis.text = element_text(size = 12),
        plot.title = element_text(size = 14),
        plot.title.position = "plot",
        plot.caption = element_text(face = "italic"),
        text = element_text(family = "Times New Roman"),
        panel.grid = element_blank())
```

**Why is this better?**  

Same points as above for the scatterplot, with some additional points:  

- title position is over the y-axis to provide some visual balance (a long title hanging over the end of the x-axis can be visually imbalanced)
- legend is taken out (because it's redundant with the x-axis)

:::{.callout-note title="Using custom colors" collapse="true"}
You can use any kind of custom colors using the `scale_color` or `scale_fill` functions.  

In this example, the colors come from the `NatParksPalettes` package. If you are using palettes from a color palette package, you will need to do some extra work to make sure you use the palette you want and generate new colors from the palette to match the number of levels in your categorical variable.  

For example, there are 12 months so you would need 12 colors. The `Glacier` palette only has 5 colors in it. Thus, the arguments would be:

```{r}
#| eval: false

NatParksPalettes::natparks.pals(name = "Glacier",     # palette name
                                n = 12,               # number of colors
                                type = "continuous")  # naming a continuous palette to force new colors to be generated
```

Then, you would have 12 additional colors:

```{r}
#| echo: false

NatParksPalettes::natparks.pals(name = "Glacier",     # palette name
                                n = 12,               # number of colors
                                type = "continuous")  # naming a continuous palette to force new colors to be generated
```

The hex codes for those colors are:

```{r}
NatParksPalettes::natparks.pals(name = "Glacier",     
                                n = 12,               
                                type = "continuous") |> 
  as.character()
```


:::



