---
title: Storing your personal access token (PAT)
description: "make sure your PAT is stored!!!"
execute: 
  eval: false
  freeze: auto
author:
  - name: An Bui
    url: https://an-bui.com/
    affiliation: UC Santa Barbara, Ecology, Evolution, and Marine Biology
    affiliation-url: https://www.eemb.ucsb.edu/
date: 2024-05-22
categories: [GitHub, Git, personal access token]
format:
  html:
    toc: true
    toc-expand: true
---

## Part 1: in RStudio console

### 1. install `usethis`

Do this in the console!

```{r}
install.packages("usethis")
```

### 2. Generate a personal access token using `usethis`

Again, do this in the console!

```{r}
usethis::create_github_token()
```

This will lead you to a web browser.

## Part 2. In the GitHub webpage

### 1. (if prompted) in the web browser, enter your GitHub password

### 2. Fill in the info about your new personal access token

### 3. Scroll to the bottom of the page, hit Generate token

### 4. Copy the personal access token to your clipboard

## Part 3. In the RStudio console

### 1. Set the personal access token.

Run this in the console!

```{r}
gitcreds::gitcreds_set()
```

### 2. Paste your personal access token when prompted

Paste your PERSONAL ACCESS TOKEN. NOT YOUR GITHUB PASSWORD. NOT YOUR COMPUTER PASSWORD.  

**PASTE YOUR PERSONAL ACCESS TOKEN.**

### 3. Double check that the personal access token is stored

Again, run this in the console:

```{r}
usethis::git_sitrep()
```

### 4. Look for `Personal access token` under `GitHub`

This should say `<discovered>`.

### 5. Look at Name and Email under `Git global (user)`

- Name should be your GitHub username  
- Email should be your GitHub account email
