---
title: "Week 5 figures - Lectures 9 and 10"
editor: source
freeze: auto
execute:
  message: false
  warning: false
format:
  html: 
    code-fold: true
author:
  - name: An Bui
    url: https://an-bui.com/
    affiliation: UC Santa Barbara, Ecology, Evolution, and Marine Biology
    affiliation-url: https://www.eemb.ucsb.edu/
published-title: "Lecture date"
date: 2024-04-29
date-modified: last-modified
categories: [analysis of variance, eta squared, shapiro-wilk, levene, mann-whitney U, wilcoxon signed-rank, kruskal-wallis, cliffs delta]
citation:
  url: https://spring-2024.envs-193ds.com/lecture/lecture_week-05.html
---

## 0. Set up

```{r set-up}
#| message: false

# cleaning
library(tidyverse)

# visualization
theme_set(theme_classic() +
            theme(panel.grid = element_blank(),
                  axis.text = element_text(size = 18),
                  axis.title = element_text(size = 18),
                  text = element_text(family = "Lato")))
library(patchwork)

# data
library(palmerpenguins)

# analysis
library(car)
library(effectsize)
```

## 1. Math

### a. sum of squares

Among groups:

$$
\sum^k_{i = 1}\sum^n_{j = 1}(\bar{y_i} - \bar{y})^2
$$

where k is the number of groups, i is ith group, n is the number of observations per group, j is the jth observation. $\bar{y_i}$ is the mean of group i, while $\bar{y}$ is the grand mean (of all samples pooled together).  

Within groups:

$$
\sum^k_{i = 1}\sum^n_{j = 1}(y_{ij} - \bar{y_i})^2
$$

where $y_{ij}$ is the jth observation in the ith group, and $\bar{y_i}$ is the mean of group i.  

Total sum of squares:  

$$
\sum^k_{i = 1}\sum^n_{j = 1}(y_{ij} - \bar{y})^2
$$

where $y_{ij}$ is the jth observation in the ith group, and $\bar{y}$ is the grand mean.

### b. mean squares

Among groups:

$$
\frac{SS_{among \ group}}{k - 1}
$$

Within groups:  

$$
\frac{SS_{within \ group}}{n - k}
$$

### c. F-ratio/F-statistic

$$
\frac{MS_{among \ group}}{MS_{within \ group}}
$$

Put another way, the F-ratio is the ratio of among group variance to within group variance. If the among group variance is larger than within group variance, the F-ratio is large, and therefore the probability of among group variance being equal to within group variance is small. Thus, you would reject the null hypothesis if the F-ratio is large.

### d. η squared

$$
\eta^2 = \frac{SS_{among \ group}}{SS_{among \ group} + SS_{within \ group}}
$$
### e. U statistic
$$
\begin{align}
U_1 &= \Sigma R_1 - n_1(n_1 + 1)/2 = 17 - 5(5+1)/2 = 2 \\
U_2 &= \Sigma R_2 - n_2(n_2 + 1)/2 = 38 - 5(5+1)/2 = 23
\end{align}
$$

## 2. Warm up: code for a figure

```{r chickwts-figure}
#| fig-width: 12
#| fig-height: 8

# random sample of 10 rows from data frame
sample_n(chickwts, 10) %>% 
  arrange(feed)

ggplot(data = chickwts,           # data frame: chickwts
       aes(x = feed,              # x-axis: feed type
           y = weight,            # y-axis: chick weight
           fill = feed)) +        # fill by feed type  
  geom_boxplot() +                # creates a boxplot
  geom_jitter(height = 0,         # prevents jitter from moving points along y-axis
              width = 0.2) +      # narrows spread of jitter along x-axis
  theme(legend.position = "none") # removes legend
```


## 3. Analysis of variance

Central question: How does bill length differ between penguin species?

### a. Exploratory data visualization

```{r penguins-explore}
penguins_jitter <- ggplot(data = penguins,
                          aes(x = species,
                              y = bill_length_mm,
                              color = species)) +
  geom_jitter(width = 0.2,
              height = 0,
              shape = 21) +
  scale_color_manual(values = c("#209c90", "#018ca9", "#27c839")) +
  labs(x = "Species",
       y = "Bill length (mm)") +
  theme(legend.position = "none")

penguins_jitter
```

### b. histogram and qq plots

```{r penguins-hist-and-qq}
#| fig-width: 12
#| fig-height: 14

hist <- ggplot(data = penguins,
       aes(x = bill_length_mm,
           fill = species)) +
  geom_histogram(bins = 14,
                 color = "black") +
  scale_fill_manual(values = c("#209c90", "#018ca9", "#27c839")) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~ species, scales = "free", ncol = 1) +
  labs(x = "Bill length (mm)",
       y = "Count") +
  theme(legend.position = "none",
        strip.background = element_rect(color = "white"),
        strip.text = element_text(size = 20))

qq <- ggplot(data = penguins,
       aes(sample = bill_length_mm)) +
  geom_qq_line() +
  geom_qq(aes(color = species)) +
  scale_color_manual(values = c("#209c90", "#018ca9", "#27c839")) +
  facet_wrap(~ species, scales = "free", ncol = 1) +
  labs(x = "Theoretical quantile",
       y = "Value") +
  theme(legend.position = "none",
        strip.background = element_rect(color = "white"),
        strip.text = element_text(size = 20))

hist + qq
```

### c. Shapiro-Wilk normality test

General: Is the response variable normally distributed?  
Example: Is bill length normally distributed?

```{r penguins-shapiro-wilk}
# first, wrangle
adelie <- penguins %>% 
  filter(species == "Adelie") %>% 
  pull(bill_length_mm)

chinstrap <- penguins %>% 
  filter(species == "Chinstrap") %>% 
  pull(bill_length_mm)

gentoo <- penguins %>% 
  filter(species == "Gentoo") %>% 
  pull(bill_length_mm)

# then, do the shapiro-wilk test
shapiro.test(adelie)
shapiro.test(chinstrap)
shapiro.test(gentoo)
```

### d. Levene test of variances

General: Are the group variances equal?  
Example: Are the species variances equal?

```{r penguins-levene}
leveneTest(bill_length_mm ~ species,
           data = penguins)
```

### e. analysis of variance

```{r penguins-anova}
# do the actual test
# model object stored as `penguins_anova`
penguins_anova <- aov(bill_length_mm ~ species,
                      data = penguins)

# output of the test
penguins_anova

# more information
summary(penguins_anova)
```

### f. post hoc test

Which group comparisons are different?

```{r penguins-tukey-hsd}
TukeyHSD(penguins_anova)
```

### g. effect size

```{r penguins-eta-squared}
eta_squared(penguins_anova)
```

### h. example of writing

Without the stats: Our results suggest a difference in bill length between species, with a large effect of species. Species differed in bill length, and pairwise comparisons between species showed that all three species differed from each other. Generally, Adelie penguins tend to have shorter bills than Chinstrap and Gentoo penguins. Gentoo penguins tend to have shorter bills than  Chinstrap penguins.

With the stats: Our results suggest a difference in bill length between species, with a large ($\eta^2$ = 0.71) effect of species. Species differed in bill length (one-way ANOVA, F(2, 339) = 410.6, p < 0.001, $\alpha$ = 0.05), and pairwise comparisons between species showed that all three species differed from each other. Generally, Adelie penguins tend to have 10.0 mm shorter bills than Chinstrap (Tukey HSD, p < 0.001, 95% confidence interval: [9.0, 11.1] mm) penguins and 8.7 mm shorter than Gentoo (Tukey HSD, p < 0.001, 95% confidence interval: [7.9, 9.6] mm) penguins. Gentoo penguins tend to have 1.3 mm shorter bills than Chinstrap penguins (Tukey HSD, p = 0.008, 95% confidence interval: [0.3, 2.4] mm).  

### i. a "finalized" figure

```{r penguins-final-fig}
ggplot(data = penguins,
       aes(x = species,
           y = bill_length_mm,
           color = species)) +
  geom_jitter(width = 0.2,
              height = 0,
              shape = 21,
              alpha = 0.4) +
  stat_summary(geom = "pointrange",
               fun.data = mean_cl_normal,
               size = 0.8,
               linewidth = 1) +
  scale_color_manual(values = c("#209c90", "#018ca9", "#27c839")) +
  labs(x = "Species",
       y = "Bill length (mm)") +
  theme(legend.position = "none")
```

## 4. Non parametric tests

### a. Mann-Whitney U

```{r wilcox-example}
wilcox_df <- cbind(Sample1 = c(1.1, 2.4, 1.8, 0.4, 1.6), 
                   Sample2 = c(5.4, 3.1, 2.3, 1.9, 4.2)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = Sample1:Sample2) %>% 
  rename(sample = name) %>% 
  arrange(sample)

wilcox_df

wilcox.test(value ~ sample,
            data = wilcox_df)
```

### b. Wilcoxon signed-rank

```{r wilcox-demo}
#| eval: false

# for a comparison of one group against a theoretical median
wilcox.test(Sample1, mu = theoretical)

# for a comparison of two groups
wilcox.test(value ~ sample,
            data = wilcox_df, 
            paired = TRUE)
```

### c. Kruskal-Wallis

```{r}
kruskal_df <- cbind(Sample1 = round(rnorm(n = 5, mean = 4, sd = 1), 1), 
                    Sample2 = round(rnorm(n = 5, mean = 6, sd = 1), 1),
                    Sample3 = round(rnorm(n = 5, mean = 8, sd = 1), 1)) %>% 
  as_tibble() %>% 
  pivot_longer(cols = Sample1:Sample3) %>% 
  rename(sample = name) %>% 
  arrange(sample)

rank_by_hand <- kruskal_df %>% 
  arrange(value) %>% 
  rownames_to_column("ranks") %>% 
  mutate(ranks = as.numeric(ranks)) %>% 
  group_by(sample) %>% 
  reframe(sum_ranks = sum(ranks))

R1 <- rank_by_hand[1, 2]
R2 <- rank_by_hand[2, 2]
R3 <- rank_by_hand[3, 2]
n <- 15
n1 <- 5
n2 <- 5
n3 <- 5

rstatix::kruskal_test(value ~ sample,
             data = kruskal_df)
kruskal.test(value ~ sample,
             data = kruskal_df)


#((12 * STATISTIC / (n * (n + 1)) - 3 * (n + 1)) / (1 - sum(TIES^3 - TIES) / (n^3 - n)))

(12/(n*(n+1)))*((R1^2)/n1 + (R2^2)/n2 + (R3^2)/n3) - 3*(n + 1)

rstatix::kruskal_effsize(value ~ sample,
                         data = kruskal_df)

rstatix::dunn_test(value ~ sample,
                         data = kruskal_df)
```




