---
title: "Lecture 06 figures"
editor: source
freeze: auto
author:
  - name: An Bui
    url: https://an-bui.com/
    affiliation: UC Santa Barbara, Ecology, Evolution, and Marine Biology
    affiliation-url: https://www.eemb.ucsb.edu/
date: 2023-05-08
categories: [chi-square, ANOVA, shapiro-wilk, levene, kruskal-wallis, cliff's delta, eta squared]
citation:
  url: https://an-bui.github.io/ES-193DS-W23/lecture/lecture-06_2023-05-08.html
---

```{r message = FALSE}
library(tidyverse)
library(palmerpenguins)
library(showtext)
library(car)
font_add_google("Lato", "Lato")
showtext_auto()
library(patchwork)
```

## Chi-square

### test statistic:

$$
\chi^2 = \Sigma\frac{(O-E)^2}{E}
$$
### degrees of freedom
$$
df = (number\;of\;rows - 1) * (number\;of\;columns - 1)
$$

### expected counts:
$$
expected = \frac{row\;total * column\;total}{table\;total}
$$

### expected counts example from lecture
$$
\frac{126 * 118}{315} = 47.2
$$

### test statistic calculation example from lecture
$$
\begin{align}
\chi^2 &= \Sigma\frac{(O-E)^2}{E} \\
\chi^2 &= \frac{55-47.2}{47.2}+...+\frac{45-31.9}{31.9} \\
&= 15.276
\end{align}
$$

### example code

```{r}
# create matrix
survey <- tribble(
  ~distance, ~trails, ~dog_access, ~wildlife_habitat,
  "walking_distance", 55, 38, 33,
  "driving_distance", 41, 25, 29,
  "out_of_town", 22, 27, 45
) %>% 
  column_to_rownames("distance")

survey

# calculate proportions
survey_summary <- tribble(
  ~distance, ~trails, ~dog_access, ~wildlife_habitat,
  "walking_distance", 55, 38, 33,
  "driving_distance", 41, 25, 29,
  "out_of_town", 22, 27, 45
) %>% 
  pivot_longer(cols = trails:wildlife_habitat, names_to = "responses", values_to = "counts") %>% 
  group_by(distance) %>% 
  mutate(sum = sum(counts)) %>% 
  ungroup() %>% 
  mutate(prop = counts/sum)
  
# do chi-square
chisq.test(survey)

# get expected matrix
chisq.test(survey)$expected
```

# ANOVA

## ANOVA variance figure:

```{r}
col1 <- "cornflowerblue"
col2 <- "orange"
col3 <- "darkgreen"

ggplot(data.frame(x = 0:22), aes(x)) +
  stat_function(geom = "line", n = 100, fun = dnorm, args = list(mean = 10, sd = 2), linewidth = 1, col = col1) + 
  stat_function(geom = "line", n = 100, fun = dnorm, args = list(mean = 8, sd = 2), linewidth = 1, col = col2) +
  stat_function(geom = "line", n = 100, fun = dnorm, args = list(mean = 15, sd = 2), linewidth = 1, col = col3) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        panel.border = element_blank(),
        text = element_text(family = "Lato")) 

```


## ANOVA with palmer penguins
```{r}
# Adelie: 10
# Chinstrap: 8
# Gentoo: 10

adelie <- penguins %>% 
  filter(species == "Adelie")

adelie_hist <- ggplot(data = adelie, aes(x = bill_length_mm)) +
  geom_histogram(bins = 10, fill = col1, color = col1, alpha = 0.8) +
  labs(x = "Bill length (mm)", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.ticks = element_blank(),
        text = element_text(family = "Lato")) 

adelie_qq <- ggplot(data = adelie, aes(sample = bill_length_mm)) +
  stat_qq_line(linewidth = 1) +
  stat_qq(col = col1) +
  labs(x = "Theoretical", y = "Sample") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.ticks = element_blank(),
        text = element_text(family = "Lato")) 

shapiro.test(adelie$bill_length_mm)

chinstrap <- penguins %>% 
  filter(species == "Chinstrap")

chinstrap_hist <- ggplot(data = chinstrap, aes(x = bill_length_mm)) +
  geom_histogram(bins = 10, fill = col2, color = col2, alpha = 0.8) +
  labs(x = "Bill length (mm)", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.ticks = element_blank(),
        text = element_text(family = "Lato")) 

chinstrap_qq <- ggplot(data = chinstrap, aes(sample = bill_length_mm)) +
  stat_qq_line(linewidth = 1) +
  stat_qq(col = col2) +
  labs(x = "Theoretical", y = "Sample") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.ticks = element_blank(),
        text = element_text(family = "Lato")) 

shapiro.test(chinstrap$bill_length_mm)

gentoo <- penguins %>% 
  filter(species == "Gentoo")

gentoo_hist <- ggplot(data = gentoo, aes(x = bill_length_mm)) +
  geom_histogram(bins = 10, fill = col3, color = col3, alpha = 0.8) +
  labs(x = "Bill length (mm)", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.ticks = element_blank(),
        text = element_text(family = "Lato")) 

gentoo_qq <- ggplot(data = gentoo, aes(sample = bill_length_mm)) +
  stat_qq_line(linewidth = 1) +
  stat_qq(col = col3) +
  labs(x = "Theoretical", y = "Sample") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        axis.ticks = element_blank(),
        text = element_text(family = "Lato")) 


(adelie_hist + adelie_qq) / (chinstrap_hist + chinstrap_qq) / (gentoo_hist + gentoo_qq)

shapiro.test(gentoo$bill_length_mm)

leveneTest(bill_length_mm ~ species, data = penguins)

penguins_anova <- aov(bill_length_mm ~ species, data = penguins)
penguins_anova
summary(penguins_anova)
TukeyHSD(penguins_anova)
```

## ANOVA information

### Sum of squares

#### among groups

$$
\sum_{i=1}^{k}\sum_{j=1}^{n}(\bar{x}_i - \bar{x})^2
$$

#### within groups
$$
\sum_{i=1}^{k}\sum_{j=1}^{n}(\bar{x}_{ij} - \bar{x}_i)^2
$$

#### total
$$
\sum_{i=1}^{k}\sum_{j=1}^{n}(\bar{x}_{ij} - \bar{x})^2
$$

### Mean squares

#### among groups

$$
\frac{SS_{among\;group}}{k-1}
$$
#### within group

$$
\frac{SS_{within\;group}}{n-k}
$$
#### total

$$
\frac{SS_{total}}{kn-1}
$$

### F-ratio

$$
\frac{MS_{among\;group}}{MS_{within\;group}}
$$

## penguins visualization
```{r}
ggplot(data = penguins, aes(x = species, y = bill_length_mm, fill = species)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2) +
  scale_fill_manual(values = c(col1, col2, col3)) +
  labs(x = "Species", y = "Bill length (mm)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        text = element_text(family = "Lato"),
        legend.position = "none") 
```


# Kruskal-Wallis

## test statistic

$$
H = \frac{12}{n(n+1)}\sum_{i = 1}^{k}\frac{R^2_i}{n_i}-3(n+1)
$$

## eta squared

$$
\eta^2 = \frac{H - k + 1}{n - k}
$$












