---
title: "Week 2 figures - Lectures 3 and 4"
editor: source
freeze: auto
execute:
  message: false
  warning: false
format:
  html: 
    code-fold: true
author:
  - name: An Bui
    url: https://an-bui.com/
    affiliation: UC Santa Barbara, Ecology, Evolution, and Marine Biology
    affiliation-url: https://www.eemb.ucsb.edu/
published-title: "Lecture date"
date: 2024-04-08
date-modified: last-modified
categories: [central limit theorem, z-distribution, t-distribution, confidence intervals, qqplot, one-tailed, two-tailed, t-test, cohen's d]
citation:
  url: https://spring-2024.envs-193ds.com/lecture/lecture_week-02.html
---

## 0. set up

```{r set-up}
#| echo: false
#| warning: false

# cleaning
library(tidyverse)
theme_set(theme_classic() +
            theme(panel.grid = element_blank(),
                  axis.text = element_text(size = 18),
                  axis.title = element_text(size = 18),
                  text = element_text(family = "Lato")))

# visualization
library(ggdist)

# panels together
library(patchwork)

# cohen's d
library(effsize)
```

## 1. Math

### a. standard error

$$
standard \: error = SE_{\bar{x}} = \frac{s}{\sqrt{n}}
$$

### b. confidence interval

$$
\begin{align}
CI = estimate \: &\pm \: margin \: of \: error \\
CI = \bar{x} \: &\pm \: t_{\alpha(2), df} \times \frac{s}{\sqrt{n}}
\end{align}
$$

### c. t-statistic

$$
t_{\alpha(2), df}
$$

$$
t_{0.05(2), 19}
$$

### d. test statistic math notation

$$
t = \frac{\bar{x_A} - \bar{x_B}}{\sqrt{\frac{s_A^2}{N_A}+\frac{s_B^2}{N_B}}}
$$

When you have unequal variances:

$$
df = \frac{(\frac{s_A^2}{N_A}+\frac{S_B^2}{N_B})^2}{\frac{(\frac{s_A^2}{N_A})^2}{N_A-1}+\frac{(\frac{s_B^2}{N_B})^2}{N_B-1}}
$$


so for two-sample t-test where you have equal variances:

$$
t = \frac{\bar{x_A} - \bar{x_B}}{s_p \times \sqrt{\frac{1}{N_A + N_B}}}
$$

the degrees of freedom are:

$$
df = (N_A - 1) + (N_B - 1)
$$

### e. Cohen's d

$$
Cohen's \; d = \frac{\bar{x_A} - \bar{x_B}}{\sqrt{(s^2_A + s^2_B)/2}}
$$

### f. confidence interval for two-sample t-test

$$
CI = (\bar{x_A} - \bar{x_B}) \pm t \times \sqrt{\frac{(n_A - 1)s_A^2 + (n_B - 1)s_B^2}{n_A+n_B-2}} \times \sqrt{\frac{1}{n_A}+\frac{1}{n_B}}
$$

## 2. Hypothesis testing: z-distribution

**Example:** You’re told that the mean weight of a coast live oak acorn is $2.1 g$. You question that claim.
You then choose to randomly sample 30 acorns and calculate the sample mean: $\bar{y} = 2.6 g$. You also (miraculously) know the population standard deviation: $\sigma = 0.3 g$.  

Do you have evidence to refute this claim? 

First, store some numbers: 
```{r acorn-values}
acorn_mean <- 2.1
acorn_sd <- 0.45
sample_mean <- 2.2
```

Plotting the acorn population distribution:

```{r acorn-pop-hist}
min <- acorn_mean-(3*acorn_sd)
max <- acorn_mean+(3*acorn_sd)

acorn_pop_hist <- data.frame(x = min:max) %>% 
  ggplot(aes(x)) +
  stat_function(geom = "line", 
                n = 1000, 
                fun = dnorm, 
                args = list(mean = acorn_mean, sd = acorn_sd), 
                linewidth = 2, 
                color = "darkgoldenrod") +
  geom_vline(xintercept = acorn_mean, 
             linetype = 2, 
             linewidth = 2) +
  scale_x_continuous(breaks = c(acorn_mean-3*acorn_sd,
                                acorn_mean-2*acorn_sd, 
                                acorn_mean-acorn_sd,
                                acorn_mean, 
                                acorn_mean+acorn_sd,
                                acorn_mean+2*acorn_sd,
                                acorn_mean+3*acorn_sd),
                     limits = c(min, max)) +
  labs(x = "Acorn mass (g)") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank())

acorn_pop_hist
```

And with the sample mean:

```{r acorn-pop-with-sample-mean}
pop_with_sample <- acorn_pop_hist +
  geom_vline(xintercept = sample_mean, 
             color = "tomato3",
             linewidth = 2)
pop_with_sample
```

And with a sample mean at least that different: 
```{r acorn-pop-with-sample-mean-at-least-that-different}
pop_with_sample +
  geom_vline(xintercept = acorn_mean - (sample_mean - acorn_mean), 
             color = "tomato3",
             linewidth = 2)
```

We can calculate a z-score using the formula:

$$
\begin{align}
z &= \frac{y - \mu}{\sigma/\sqrt{n}} \\
&= \frac{2.2 - 2.10}{0.45/\sqrt{30}} \\
&= 1.22
\end{align}
$$

In this version of the z-score formula, the number of observations $n$ is included; technically, the one from before also included $n$, but since we were only choosing one individual and asking about the probability of selecting that individual, $\sqrt{n} = 1$ so it cancelled out.

```{r acorn-z}
acorn_z <- (sample_mean - acorn_mean)/(acorn_sd/sqrt(30))
acorn_z

pnorm(acorn_z, mean = 0, sd = 1, lower.tail = FALSE)
```

We can plot the z-scores against each other:
```{r plotting-z}
data.frame(x = -3:3) %>% 
  ggplot(aes(x)) +
  stat_function(geom = "area",
                fun = dnorm,
                args = list(mean = 0, sd = 1),
                xlim = c(1.96, 3),
                fill = "lightgrey") +
  stat_function(geom = "area",
                fun = dnorm,
                args = list(mean = 0, sd = 1),
                xlim = c(-3, -1.96),
                fill = "lightgrey") +
  geom_linerange(x = 1.96, ymin = 0, ymax = 0.06) +
  geom_linerange(x = -1.96, ymin = 0, ymax = 0.06) +
  geom_linerange(x = acorn_z, linetype = 2, linewidth = 2,
                 ymin = 0, ymax = 0.18) +
  geom_linerange(x = -acorn_z, linetype = 2, linewidth = 2,
                 ymin = 0, ymax = 0.18) +
  stat_function(geom = "line", 
                n = 1000, 
                fun = dnorm, 
                args = list(mean = 0, sd = 1), 
                linewidth = 2, 
                color = "darkgoldenrod") +
  scale_x_continuous(breaks = seq(-3, 3, by = 1)) +
  scale_y_continuous(expand = c(0, 0),
                     limits = c(0, 0.45)) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank()) +
  labs(x = "Z")
```


## 2. central limit theorem

If you were to sample a bunch of times from any distribution (i.e. take many observations within a sample, take many observations in another sample), the mean values for each sample will be normally distributed. Kareem Carr has a nice explainer of how this works [here](https://twitter.com/kareem_carr/status/1754524686606626894).

```{r clt-setup-uniform-distribution}
#| fig-width: 12
#| fig-height: 8

# randomly select 10000 numbers from a uniform distribution for the population
uniform <- runif(10000, min = 2, max = 8)

# make a histogram for the population
uniformdf <- as.data.frame(uniform)

ggplot(uniformdf, aes(x = uniform)) +
  geom_histogram(breaks = seq(2, 8, length.out = 41), fill = "firebrick", alpha = 0.7, color = "firebrick") +
  geom_vline(xintercept = mean(uniform), linewidth = 2) +
  annotate("text", x = 4, y = 290, label = "mean = 4.967", size = 10) +
  scale_x_continuous(breaks = seq(from = 2, to = 8, by = 1)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 305)) +
  labs(x = "Continuous value", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18))
```


```{r clt-samples}
#| fig-width: 12
#| fig-height: 8

# for() loop to 
store2 <- c()
store5 <- c()
store15 <- c()
store30 <- c()
store50 <- c()

for(i in 1:100) {
  
  # sample 100x
  store2[i] <- mean(sample(uniform, 2, replace = FALSE))
  store5[i] <- mean(sample(uniform, 5, replace = FALSE))
  store15[i] <- mean(sample(uniform, 15, replace = FALSE))
  store30[i] <- mean(sample(uniform, 30, replace = FALSE))
  store50[i] <- mean(sample(uniform, 50, replace = FALSE))

}


df <- cbind(store2, store5, store15, store30, store50) %>% 
  as.data.frame()
  
ggplot(df) +
  geom_histogram(aes(x = store2), bins = 10, alpha = 0.7, fill = "chocolate1", color = "chocolate1") +
  coord_cartesian(xlim = c(2, 8), ylim = c(0, 30)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_vline(xintercept = mean(store2)) +
  geom_vline(xintercept = mean(uniform), color = "red") +
  labs(x = "Sample means", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm"))


ggplot(df) +
  geom_histogram(aes(x = store5), bins = 10, alpha = 0.7, fill = "blue3", color = "blue3") +
  coord_cartesian(xlim = c(2, 8), ylim = c(0, 30)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_vline(xintercept = mean(store5)) +
  geom_vline(xintercept = mean(uniform), color = "red") +
  labs(x = "Sample means", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm"))

ggplot(df) +
  geom_histogram(aes(x = store15), bins = 12, alpha = 0.7, fill = "darkorchid4", color = "darkorchid4") +
  coord_cartesian(xlim = c(2, 8), ylim = c(0, 30)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_vline(xintercept = mean(store15)) +
  geom_vline(xintercept = mean(uniform), color = "red") +
  labs(x = "Sample means", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm"))

ggplot(df) +
  geom_histogram(aes(x = store30), bins = 12, alpha = 0.7, fill = "lightseagreen", color = "lightseagreen") +
  coord_cartesian(xlim = c(2, 8), ylim = c(0, 30)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_vline(xintercept = mean(store30)) +
  geom_vline(xintercept = mean(uniform), color = "red") +
  labs(x = "Sample means", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm"))

ggplot(df) +
  geom_histogram(aes(x = store50), bins = 12, alpha = 0.7, fill = "violetred3", color = "violetred3") +
  coord_cartesian(xlim = c(2, 8), ylim = c(0, 30)) +
  scale_y_continuous(expand = c(0, 0)) +
  geom_vline(xintercept = mean(store50)) +
  geom_vline(xintercept = mean(uniform), color = "red") +
  labs(x = "Sample means", y = "Count") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        plot.margin = unit(c(0.5, 0.5, 0.1, 0.1), "cm"))

```

## 3. z- vs t-distribution

### a. comparison with normal

t-distributions allow for more uncertainty around the tails.

```{r comparison-plot}
#| fig-width: 12
#| fig-height: 8

ggplot(data.frame(x = -5:5), aes(x)) +
  stat_function(geom = "line", n = 1000, fun = dnorm, args = list(mean = 0, sd = 1), linewidth = 1, color = "darkorange") +
  annotate("text", x = 2.5, y = 0.4, label = "normal", color = "darkorange", size = 6) +
  stat_function(geom = "line", n = 1000, fun = dt, args = list(df = 1), linewidth = 1, color = "#856F33") +
  annotate("text", x = 3, y = 0.32, label = "t-distribution (small n)", color = "#856F33", size = 6) +
  stat_function(geom = "line", n = 1000, fun = dt, args = list(df = 10), linewidth = 1, color = "#56E9E7") +
  annotate("text", x = 3, y = 0.37, label = "t-distribution (large n)", color = "#56E9E7", size = 6) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, 0.42)) +
  labs(x = "Continuous value", y = "Density") +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18),
        text = element_text(family = "Lato")) 
```

### b. visual representation of significance and t-statistic

```{r t-dist-and-significance}
#| fig-width: 14
#| fig-height: 6
#| fig-align: center

ggplot(data.frame(x = -5:5), aes(x)) +
  stat_function(geom = "area", fun = dt, args = list(df = 19), xlim = c(1.8, 5), fill = "#0070c0") +
  stat_function(geom = "area", fun = dt, args = list(df = 19), xlim = c(-5, -1.8), fill = "#0070c0") +
  stat_function(geom = "line", n = 1000, fun = dt, args = list(df = 19), linewidth = 1, color = "#000000") +
  geom_hline(yintercept = 0) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.4)) +
  theme_void() +
  theme(panel.grid = element_blank(),
        plot.margin = unit(c(1, 0, 0, 0), "cm"))
```

## 4. confidence intervals

This is the leaf example from lecture.

### random number generation

This generates the "population": 10000 trees.

```{r leaf-number-generation}
set.seed(7)
leaf_pop <- rnorm(n = 10000, mean = 4.92, sd = 0.5)
leaves <- sample(leaf_pop, size = 20, replace = FALSE)
```

### population histogram

```{r leaf-pop-hist}
# population histogram
enframe(leaf_pop) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(fill = "darkgreen", 
                 color = "darkgreen",
                 alpha = 0.8) +
  geom_vline(xintercept = mean(leaf_pop),
             linetype = 2,
             linewidth = 2) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Leaf length (cm)",
       y = "Count") +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank())
```

### sample histogram

```{r leaf-histogram}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center
breakpoints <- round(seq(from = min(leaves), to = max(leaves), length.out = 7), 2)

hist <- enframe(leaves) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(bins = 7, fill = "cornflowerblue", color = "#000000", breaks = breakpoints) +
  scale_x_continuous(breaks = breakpoints) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 6), breaks = c(0, 1, 2, 3, 4, 5, 6)) +
  theme_classic() +
  labs(x = "Leaf length (cm)", y = "Count")
hist
```

### sample density plot
```{r sample-density}
enframe(leaves) %>% 
  ggplot(aes(x = value)) +
  geom_density(fill = "cornflowerblue", color = "#000000", linewidth = 1) +
  scale_x_continuous(limits = c(2.5, 7)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.65)) +
  geom_vline(xintercept = mean(leaves)) +
  annotate("text", x = 6, y = 0.6, label = "sample mean = 4.84 cm", size = 6) +
  theme_classic() +
  labs(x = "Leaf length (cm)", 
       y = "Density")
```

### sample dot and whisker plot with confidence intervals
```{r sample-dot-and-whisker}
leaf_conflev <- tribble(
  ~ conflev,
  0.85,
  0.90,
  0.95,
  0.99
) %>% 
  mutate(lower = case_when(
    conflev == 0.85 ~ mean(leaves) - -qt(p = 0.15/2, df = 19)*sd(leaves)/sqrt(length(leaves)),
    conflev == 0.90 ~ mean(leaves) - -qt(p = 0.1/2, df = 19)*sd(leaves)/sqrt(length(leaves)),
    conflev == 0.95 ~ mean(leaves) - -qt(p = 0.05/2, df = 19)*sd(leaves)/sqrt(length(leaves)),
    conflev == 0.99 ~ mean(leaves) - -qt(p = 0.01/2, df = 19)*sd(leaves)/sqrt(length(leaves)) 
  ), 
  upper = case_when(
    conflev == 0.85 ~ mean(leaves) + -qt(p = 0.15/2, df = 19)*sd(leaves)/sqrt(length(leaves)),
    conflev == 0.90 ~ mean(leaves) + -qt(p = 0.1/2, df = 19)*sd(leaves)/sqrt(length(leaves)),
    conflev == 0.95 ~ mean(leaves) + -qt(p = 0.05/2, df = 19)*sd(leaves)/sqrt(length(leaves)),
    conflev == 0.99 ~ mean(leaves) + -qt(p = 0.01/2, df = 19)*sd(leaves)/sqrt(length(leaves))
  )) %>% 
  mutate(mean = mean(leaves))
  # se <- s/sqrt(n)

ggplot() +
  geom_point(data = enframe(leaves), aes(x = 0.84, y = leaves),
             alpha = 0.6, shape = 21) +
  geom_point(data = enframe(leaves), aes(x = 0.89, y = leaves),
             alpha = 0.6, shape = 21) +
  geom_point(data = enframe(leaves), aes(x = 0.94, y = leaves),
             alpha = 0.6, shape = 21) +
  geom_point(data = enframe(leaves), aes(x = 0.98, y = leaves),
             alpha = 0.6, shape = 21) +
  geom_point(data = leaf_conflev, aes(x = conflev, y = mean), 
             size = 3,
             color = "cornflowerblue") +
  geom_errorbar(data = leaf_conflev, aes(x = conflev, y = mean, ymin = lower, ymax = upper), 
                width = 0.006, 
                linewidth = 1,
                color = "cornflowerblue") +
  theme_void() +
  theme(panel.grid = element_blank()) +
  labs(x = "Confidence levels", 
       y = "Leaf length (cm)") 


```

### sample qq plot
```{r leaf-qq-plot}
#| fig-width: 8
#| fig-height: 6
#| fig-align: center

qq <- enframe(leaves) %>% 
  ggplot(aes(sample = value)) +
  stat_qq_line(aes(sample = value)) +
  stat_qq(aes(sample = value), color = "cornflowerblue", size = 3) +
  theme_classic() +
  labs(x = "Theoretical quantiles", y = "Sample quantiles")
```

```{r sample-histogram-and-qq-plot}
hist + qq
```

### resampling visual

#### resampling with different sample sizes

```{r sampling-distributions}
#| fig.width: 6
#| fig-height: 8

leaf_5 <- rep(NA, length = 1000)
leaf_20 <- rep(NA, length = 1000)
leaf_40 <- rep(NA, length = 1000)

leaf_20_sd <- rep(NA, length = 1000)

# sample 5 leaves from population 1000x
for(i in 1:1000) {
  
  # sample 5 leaves from population
  sample_5 <- sample(leaf_pop, size = 5, replace = FALSE) 
  sample_20 <- sample(leaf_pop, size = 20, replace = FALSE) 
  sample_40 <- sample(leaf_pop, size = 40, replace = FALSE) 
  
  leaf_5[i] <- mean(sample_5)
  leaf_20[i] <- mean(sample_20)
  leaf_40[i] <- mean(sample_40)
  
  leaf_20_sd[i] <- sd(sample_20)
}

leaf_df <- cbind(leaf_5, leaf_20, leaf_40) %>% 
  as_tibble() %>% 
  pivot_longer(cols = 1:3) %>% 
  mutate(name = case_when(
    name == "leaf_5" ~ "n = 5",
    name == "leaf_20" ~ "n = 20",
    name == "leaf_40" ~ "n = 40"
  ),
  name = fct_relevel(name, "n = 5", "n = 20", "n = 40"))

leaf_df %>% 
  filter(name == "n = 20") %>% 
  ggplot() +
  geom_histogram(aes(x = value),
                 color = "cornflowerblue",
                 fill = "cornflowerblue",
                 alpha = 0.8) +
  geom_vline(xintercept = mean(leaf_pop),
             linetype = 2,
             linewidth = 2) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 280)) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        strip.text = element_text(size = 18)) +
  labs(x = "Leaf length (cm)") +
  facet_wrap(~name, ncol = 1) 

ggplot(leaf_df) +
  geom_histogram(aes(x = value),
                 color = "cornflowerblue",
                 fill = "cornflowerblue",
                 alpha = 0.8) +
  geom_vline(xintercept = mean(leaf_pop),
             linetype = 2,
             linewidth = 2) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 280)) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        strip.text = element_text(size = 18)) +
  labs(x = "Leaf length (cm)") +
  facet_wrap(~name, ncol = 1) 
```

#### resampling confidence intervals

```{r resampling-n-20}
#| fig-width: 6
#| fig-height: 8

leaf_20_ci <- cbind(leaf_20, leaf_20_sd) %>% 
  as_tibble() %>% 
  mutate(ci_low = leaf_20 - -qt(p = 0.05/2, df = 19)*leaf_20_sd/sqrt(20),
         ci_high = leaf_20 + -qt(p = 0.05/2, df = 19)*leaf_20_sd/sqrt(20),
         iter = rownames(.)) %>% 
  mutate(color = case_when(
    ci_low <= mean(leaf_pop) & ci_high >= mean(leaf_pop) ~ "yes",
    TRUE ~ "no"
  ),
  color = fct_relevel(color, "yes", "no"))

# selecting 8 resamples to plot
leaf_20_ci_sample <- leaf_20_ci %>% 
  group_by(color) %>% 
  sample_n(4) %>% 
  ungroup() %>% 
  mutate(iter = fct_inorder(iter))

leaf_20_ci_sample

ggplot(data = leaf_20_ci_sample, aes(x = leaf_20, y = iter, color = color)) +
  geom_vline(xintercept = mean(leaf_pop),
             lty = 2) +
  geom_pointrange(aes(xmin = ci_low, xmax = ci_high)) +
  scale_color_manual(values = c("yes" = "darkgreen", "no" = "orange")) +
  scale_x_continuous(limits = c(mean(leaf_pop)*0.85, mean(leaf_pop)*1.15)) +
  scale_y_discrete(limits = rev) +
  theme(axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  theme(legend.position = "none") +
  labs(x = "Leaf length (cm)")
```

### plotting standard error

```{r standard-error}
#| fig-width: 4
#| fig-height: 6
enframe(leaves) %>% 
  mutate(group = "Sample") %>% 
  ggplot(aes(x = group, y = value)) +
  geom_point(position = position_jitter(width = 0.2, height = 0, seed = 1),
             shape = 21,
             alpha = 0.8) +
  geom_point(aes(x = group, y = mean(value)),
             color = "cornflowerblue",
             size = 4) +
  geom_errorbar(aes(ymin = mean(value) - sd(value)/sqrt(20), 
                    ymax = mean(value) + sd(value)/sqrt(20),
                    width = 0.2),
                color = "cornflowerblue",
                linewidth = 1) +
  labs(y = "Leaf length (cm)") +
  theme(axis.title.x = element_blank())
```

### plotting with "dry" trees

```{r dry-comparison-plot}
enframe(leaves) %>% 
  mutate(group = "Sample") %>% 
  ggplot() +
  geom_point(aes(x = group, y = value),
             position = position_jitter(width = 0.2, height = 0, seed = 1),
             shape = 21,
             alpha = 0.8) +
  geom_point(aes(x = group, y = mean(value)),
             color = "cornflowerblue",
             size = 3) +
  geom_errorbar(data = leaf_conflev %>% filter(conflev == 0.95) %>% mutate(group = "Sample"),
                aes(x = group, ymin = lower, ymax = upper),
                width = 0,
                color = "cornflowerblue",
                size = 1) +
  geom_pointrange(data = data.frame(group = "Dry", mean = 3.2, lower = 3.0, upper = 3.4),
                  aes(x = group, y = mean, ymin = lower, ymax = upper),
                  color = "firebrick",
                  size = 1,
                  linewidth = 1) +
  labs(y = "Leaf length (cm)") +
  theme(axis.title.x = element_blank())
```


## 5. qqplot examples

We use qqplots (quantile-quantile plots) to visually evaluate the normality of some variable. The x-axis is the "theoretical" quantile, and the y-axis is the "sample" quantile. If the points follow a 1:1 line, then the variable is normally distributed.  

The New Haven temperature data is normally distributed:

```{r nhtemp-histogram-and-qq}
#| fig-width: 16
#| fig-height: 6

nhtemp_hist <- as_tibble(nhtemp) %>% 
  ggplot(aes(x = x)) +
  geom_histogram(breaks = seq(47, 55, length.out = 9), fill = "turquoise3", color = "#000000") +
  scale_x_continuous(breaks = seq(47, 55, length.out = 9), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 23)) +
  theme_classic() +
  labs(x = "Bins", y = "Count") +
  theme(plot.margin = unit(c(0.1, 1, 0.1, 0.1), "cm")) 

nhtemp_qq <- ggplot(as_tibble(nhtemp)) +
  stat_qq(aes(sample = x), color = "turquoise3", size = 3) +
  labs(x = "Theoretical quantile", y = "Sample quantile") +
  theme(plot.margin = unit(c(0.1, 1, 0.1, 0.1), "cm")) 

nhtemp_hist + nhtemp_qq
```

The sunspot data is not:

```{r sunspots-histogram-and-qq}
#| fig-width: 10
#| fig-height: 6

sunspot_hist <- as_tibble(sunspots) %>% 
  ggplot(aes(x = x)) +
  geom_histogram(breaks = round(seq(0, 260, length.out = 30)), fill = "tomato2", color = "#000000") +
  scale_x_continuous(breaks = round(seq(0, 260, length.out = 30)), expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 480)) +
  theme_classic() +
  labs(x = "Bins", y = "Count") +
  theme(plot.margin = unit(c(0.1, 1, 0.1, 0.1), "cm")) 

sunspot_qq <- ggplot(as_tibble(sunspots)) +
  stat_qq(aes(sample = x), color = "tomato2", size = 3) +
  theme_classic() +
  labs(x = "Theoretical quantile", y = "Sample quantile") +
  theme(plot.margin = unit(c(0.1, 1, 0.1, 0.1), "cm")) 

sunspot_hist 
sunspot_qq

```

