---
title: "OPTIONAL practice problem - Do a Wilcoxon signed rank test by hand"
editor: source
freeze: auto
date: 2024-05-01
date-modified: last-modified
---

```{r document-set-up}
#| echo: false
#| message: false

# package
library(tidyverse)

theme_set(theme_bw() +
            theme(text = element_text(family = "Lato", size = 12)))
```

## 1. Description

In this practice problem, you'll do a Wilcoxon signed rank test comparing paired samples to each other by hand, then double check your work with R. This is the non-parametric version of the paired t-test.  

## 2. Steps

### a. Values

Start with these values:  

:::: {.columns}
::: {.column width="40%"}

| Before | After |
|:------:|:-----:|
| 1.2    | 1.6   |
| 2.0    | 2.8   |
| 0.8    | 0.7   |
| 0.6    | 0.6   |
| 1.3    | 1.5   |
| 3.2    | 2.7   |
| 0.9    | 1.1   |

:::
::::

### b. Calculate the differences between samples

Then, calculate the differences between those values. You should get something that looks like this:

:::: {.columns}
::: {.column width="40%"}

| Differences    |
|:--------------:|
| -0.4           |
| -0.8           |
| 0.1            |
| 0.0            |
| -0.2           |
| 0.5            |
| -0.2           |

:::
::::

### c. Take out 0, arrange by magnitude

Then, omit all observations of 0 and arrange your numbers by magnitude. You should get something that looks like this:

:::: {.columns}
::: {.column width="40%"}

| Ordered magnitude    |
|:--------------------:|
| 0.1                  |
| -0.2                 |
| -0.2                 |
| -0.4                 |
| 0.5                  |
| -0.8                 |

:::
::::

### d. Give each value a "sign" and a "rank"

This is the "signed rank" part of the test: assign each value a sign (+ or -), then rank (1, 2, 3, 4, etc.) You should get something that looks like this:

:::: {.columns}
::: {.column width="40%"}

| Ordered magnitude    |   Signed rank        |
|:--------------------:|:--------------------:|
| 0.1                  |      + 1             |
| -0.2                 |      - 2             |
| -0.2                 |      - 2             |
| -0.4                 |      - 4             |
| 0.5                  |      + 5             |
| -0.8                 |      - 6             |

:::
::::

### e. Sum the magnitudes of the + and - values

Sum the magnitudes of the + and - values. This will be the -W and +W statistic. You then choose the lower statistic.

In this case, you should get:  

+W = 1 + 5 = 6  
-W = 2 + 2 + 4 + 6 = 14  

Pick the lowest one: +W = 6.  

### e. Run the test in R to check your work.

Start with this data frame:

```{r}
# creating a long format data frame
df <- tibble(before = c(1.2, 2.0, 0.8, 0.6, 1.3, 3.2, 0.9), # original values
             after = c(1.6, 2.8, 0.7, 0.6, 1.5, 2.7, 1.1)) %>% 
  pivot_longer(cols = before:after, # making this long format
               names_to = "sample", # creating a column called "sample" that has before and after
               values_to = "value") %>% # creating a column called value that has the values
  mutate(sample = fct_relevel(sample, "before", "after")) %>% # making sure that before appears in the data frame first
  arrange(sample) # arranging the data frame so that before and after are all sitting 
```

Then run a Wilcoxon signed rank test (note the `paired = TRUE` argument).

```{r}
wilcox.test(value ~ sample, # formula
            data = df, # data
            paired = TRUE) # argument for a paired test
```

### f. Try a one-sample test.

Compare the following values to $\mu = 7$:

:::: {.columns}
::: {.column width="40%"}

| Sample |
|:------:|
| 2.8    |
| 8.3    |
| 4.5    |
| 6.7    |
| 8.1    |
| 7.2    |
| 7.8    |

:::
::::


First, calculate the difference between each observation and the $\mu$, which is 7:

:::: {.columns}
::: {.column width="40%"}

| Sample |  Difference  |
|:------:|:------------:|
| 2.8    |  -4.2        |
| 8.3    |  1.3         |
| 4.5    |  -2.5        |
| 6.7    |  -0.3        |
| 8.1    |  1.1         |
| 7.2    |  0.2         |
| 7.8    |  0.8         |

:::
::::


Then order each difference and assign a sign and rank:

:::: {.columns}
::: {.column width="40%"}

| Ordered magnitude    |   Signed rank        |
|:--------------------:|:--------------------:|
| 0.2                  |      + 1             |
| -0.3                 |      - 2             |
| 0.8                  |      + 3             |
| 1.1                  |      + 4             |
| 1.3                  |      + 5             |
| -2.5                 |      - 6             |
| -4.2                 |      - 7             |

:::
::::


Then sum the magnitudes:  

+W = 1 + 3 + 4 + 5 = 13   
-W = 2 + 6 + 7 = 15  

For this example, W = 13.  

Then, double check your work in R:

```{r}
wilcox.test(c(2.8, 8.3, 4.5, 6.7, 8.1, 7.2, 7.8),
            mu = 7)
```



