---
title: "OPTIONAL practice problem - Make an ugly plot"
editor: source
freeze: auto
date: 2024-04-25
date-modified: last-modified
---

# Why make an ugly plot?

Making ugly things is a good way of getting over the barrier of getting to something good - sometimes you just need to get something down on paper (or code, as it were) to get started.  

The only issue is that there are lots of little things to tweak in `ggplot()` or any other code to make a visualization, and it can be hard to parse all the little components when you're trying to make a good figure you can put in a presentation or report or paper.  

However, if you've made an ugly plot, you will likely have already learned about all the components in a `ggplot` object that you can control. 

# Problem

Take the visualization you made for Workshop 4 and make it ugly. If you feel compelled to do so, post it on the discussion forum on Canvas so that everyone can marvel at your ugly plot!  

# Starting out

This is a figure using the Trash Wheel data set from Workshop 4. 

```{r set-up}
#| message: false

# read in packages
library(tidyverse)

# read in data
trashwheel <- read_csv(here::here("workshop", "data", "trashwheel.csv"))
```

I'm going to start with a basic plot with date on the x-axis and weight on the y-axis. I'll also create lines and points representing the trash collected by each trash wheel.

```{r starter-plot}
trash_plot <- ggplot(data = trashwheel,      # use the trash wheel data frame
                     aes(x = date,           # put date on the x-axis
                         y = weight,         # put weight on the y-axis
                         color = name)) +    # color everything by the name of the trash wheel
  geom_point() +                             # make a layer of points
  geom_line() +                              # make a layer of lines
  theme(legend.position = "bottom") +        # put the legend on the bottom
  facet_wrap(~ name)                         # separate the trash wheels into different columns

trash_plot
```

From this visualization, it seems that the original Mister Trash Wheel collects more trash in weight than the other trash wheels. What are the units of weight? Who knows!

# Making things worse

You can play around with the `theme()` arguments to really make your ugly plot shine. If you add each argument one by one (for example, I added `axis.title.x`, looked at the output, then added `axis.title.y` and so forth), you can start to understand which argument goes with what part of the plot. 

```{r uglier-plot}
trash_plot +
  theme(
    
    # axis adjustments
    axis.title.x = element_text(angle = 180, family = "Papyrus"),
    axis.title.y = element_text(angle = 60, family = "Times New Roman"),
    axis.text.x = element_text(angle = 120, family = "Garamond"),
    axis.text.y = element_text(color = "lightblue", angle = 180, family = "Bradley Hand"),
    axis.ticks.y = element_line(color = "red"),
    axis.ticks.x.bottom = element_line(color = "limegreen"),
    axis.line.y = element_line(color = "purple"),
    
    # legend adjustments
    legend.background = element_rect(fill = "blue"),
    legend.direction = "vertical",
    legend.title = element_text(color = "white", family = "American Typewriter"),
    legend.text = element_text(color = "pink", family = "Curlz MT"),
    
    # panel adjustments
    panel.background = element_rect(color = "orange", fill = "darkred"),
    panel.grid.major.x = element_line(linetype = 3),
    panel.grid.minor.x = element_line(linetype = 6, color = "purple4"),
    panel.grid.minor.y = element_line(linetype = 5, color = "orchid"),
    
    # plot (as in, everything that's not the panel)
    plot.background = element_rect(fill = "#289793"),
    
    # strips for the facet
    strip.background = element_rect(fill = "#809283"),
    strip.text = element_text(size = 14, family = "Imprint MT Shadow")
    
  )
```

Take the plot you made in workshop and make it ugly. If you want to share your plot, post it on the discussion forum in Canvas!






