---
title: "Problem set 4"
editor: source
published-title: "Due date"
date: 2026-05-21
date-modified: last-modified
categories: [problem set]
execute:
  message: false
  warning: false
---

## Set up

```{r}
#| echo: false

library(tidyverse)
library(lterdatasampler)

theme_set(
  theme_bw() +
    theme(text = element_text(size = 12))
)
```


You may turn in your problem set in .pdf or .docx format. This can be generated using code in a .qmd or .Rmd file, or a regular Word or Google Docs document.    

A **complete** problem set will include:  

- your name and the date
- the title ("Problem set 4")
- answers to each problem component in the listed order (e.g. 1a, 1b, 1c, and so on)
- all answers _visible_ with no missing components

Additionally, a **complete** problem set _must_ be accompanied by a Generative AI statement of use (linked in the portal on Canvas).  

Any problem set missing these components will be marked **incomplete**. See the syllabus for information on how missing problem sets) factor into your course grade.

## Problems

### Problem 1. Water temperature at different latitudes

Latitude typically influences temperature (measured in °C).

Based on the plot below, do you think there is a negative, positive, or no relationship between latitude and air temperature? Do you think there is a linear relationship between air temperature and latitude? Be specific about the visual components (i.e. the points) that allow you to make this claim.

```{r}
#| echo: false

pie_crab_clean <- pie_crab |> 
  select(latitude, air_temp) |> 
  distinct()

ggplot(data = pie_crab_clean,
       mapping = aes(x = latitude,
                     y = air_temp)) +
  geom_point(size = 2,
             color = "darkblue",
             shape = 21) +
  labs(x = "Latitude",
       y = "Air temperature (\U00B0 C)")
```


### Problem 2. Evaluating diagnostic plots

You've fit a linear model to answer your question in problem 2.

```{r}
#| echo: false

par(mfrow = c(2, 2))

lm(air_temp ~ latitude,
   data = pie_crab_clean) |> 
  plot()
```

a. What do residuals vs fitted and scale-location plots tell you? What patterns are you looking for?
b. What do QQ Residuals plots tell you? What patterns are you looking for?
c. What do Residuals vs Leverage plots tell you? What patterns are you looking for?
d. Based on these diagnostics, do you think that a linear model was the appropriate analysis to examine the predictive relationship between latitude and temperature?

### Problem 3. Trout weights and lengths

You're interested in the relationship between weight and length of cutthroat trout. Based on the figure below, do you think there is a negative, positive, or no relationship between length and weight? Do you think there is a linear relationship between length and weight? Be specific about the visual compoments of the plot (i.e. the points) that you are using to make this claim.

```{r}
#| echo: false


ggplot(data = and_vertebrates |> filter(species == "Cutthroat trout" & section == "OG"),
       mapping = aes(x = length_1_mm,
                     y = weight_g)) +
  geom_point(shape = 21,
             alpha = 0.7,
             color = "darkblue") +
  labs(x = "Length (mm)",
       y = "Weight (g)")
```

### Problem 4. Different diagnostics

You fit a model to answer your question in problem 3.

```{r}
#| echo: false

par(mfrow = c(2, 2))

lm(weight_g ~ length_1_mm,
   data = and_vertebrates |> filter(species == "Cutthroat trout" & section == "OG")) |> 
  plot()
```

a. What would you say about the patterns in the Residuals vs Fitted and the Scale-Location plots?
b. What would you say about the patterns in the QQ Residuals plots?
c. Based on these diagnostics, do you think that a linear model was the appropriate analysis to examine the predictive relationship between length and weight of cutthroat trout?

### Problem 5. Relationships

You're interested in the relationship between water elevation and conductivity in a stream ecosystem. You make a plot that looks like this:

:::{.column-margin}
edited 2026-05-19: clarifying figure and variables
:::

```{r}
#| echo: false

ggplot(data = luq_streamchem,
       mapping = aes(x = gage_ht,
                     y = cond)) +
  geom_point() +
  labs(x = "Water elevation (m)",
       y = expression("Conductivity (\U03BC S/cm)"))
```


You're not sure which test to run, so you try both.

```{r}

cor.test(luq_streamchem$gage_ht, luq_streamchem$cond, method = "pearson")

cor.test(luq_streamchem$gage_ht, luq_streamchem$cond, method = "spearman")
```

a. What are the two types of correlation?
b. Based on these outputs, what would you decide about your null hypothesis? Would your decision differ based on the test you ran?
c. Choose one output to write about. Be sure to include all the components of the test summary to transparently communicate about the mechanics of the test.














