---
title: "Problem set 2"
editor: source
published-title: "Due date"
date: 2026-04-23
date-modified: last-modified
categories: [problem set]
---

# Set up

Refer back to the directions for Problem set 1 in the "Set up" section for formatting and completion guidelines.

```{r}
#| echo: false
#| message: false

library(tidyverse)
library(lterdatasampler)
```

# Problems

## 1. Spread vs precision

a. Mathematically, how do SD and SE differ (in their formulas)?  

b. Conceptually, how do SD and SE differ (i.e. how do they differ in what they represent)?  

## 2. Hypothesis writing

What are the main characteristics of a:  

a. biological,

b. statistical, and

c. mathematical hypothesis?

## 3. Hypothesis testing

### a. Pickleweed

You read in a paper that mean pickleweed (_Salicornia virginica_) stem diameter in a slough in San Diego is 0.43 cm. You think that the pickleweed at Devereux Slough has a different height.

- What are your null and alternative hypotheses in biological _and_ statistical terms?
- What test would you choose to test your hypothesis? Justify your answer.

### b. Oak growth rates

You think that growth rate (measured in cm/week) of oak trees grown in the greenhouses will increase after a few weeks of nitrogen addition to the soil in which they are growing. You measure growth rate for two weeks before adding nitrogen, then measure growth rate for two weeks after adding nitrogen.

- What are your null and alternative hypotheses in biological _and_ statistical terms?
- What test would you choose to test your hypotheses? Justify your answer.

## 4. Interpreting p-values

### a. Mean monthly air temperature

You're comparing _mean_ daily air temperature between January and March at Toolik Field Station. 

This is some summary information about the data:

```{r}
#| echo: false

temp_df <- arc_weather |> 
  mutate(month = month(date, label = TRUE),
         year = year(date)) |>
  filter(year == 2018 & month %in% c("Jan", "Mar")) |> 
  rename(mean_daily_airtemp = mean_airtemp)

temp_df |> 
  group_by(month) |> 
  summarize(mean_monthly_airtemp = mean(mean_daily_airtemp),
            n = length(mean_daily_airtemp)) |> 
  ungroup()
```


And this is the result of your test:

```{r}
#| echo: false

t.test(
  mean_daily_airtemp ~ month,
  data = temp_df,
  var.equal = FALSE
)
```

- What kind of t-test is this?
- What _must_ be true to run this type of t-test?
- Write a statement summarizing the results of this test in 1-2 sentences.

### b. Crab carapace size

You're comparing the carapace size (measured in mm) of crabs between Narragansett Bay National Estuarine Research Reserve and Cape Cod. 

This is some summary information about the data:

```{r}
#| echo: false

crab_df <- pie_crab |> 
  filter(site %in% c("CC", "NB"))

crab_df |> 
  group_by(name) |> 
  summarize(mean_size = mean(size),
            n = length(size)) |> 
  ungroup()
```


This is the result of your test:

```{r}
#| echo: false

t.test(
  size ~ name,
  data = crab_df,
  var.equal = TRUE
)
```

- What kind of t-test is this?
- What _must_ be true to run this type of t-test?
- Write a statement summarizing the results of this test in 1-2 sentences.

### c. Trout weights

You're comparing the weights of cutthroat trout (measured in g) between cascade vs pool habitats.

This is some summary information about the data:

```{r}
#| echo: false

trout_df <- and_vertebrates |> 
  filter(species == "Cutthroat trout" & unittype %in% c("C", "P")) |> 
  mutate(unittype = case_when(
    unittype == "C" ~ "cascade",
    unittype == "P" ~ "pool"
  ))

trout_df |> 
  group_by(unittype) |> 
  summarize(mean_weight = mean(weight_g, na.rm = TRUE),
            n = length(weight_g)) |> 
  ungroup()
```

This is the result of your test:

```{r}
#| echo: false

t.test(
  weight_g ~ unittype,
  data = trout_df,
  var.equal = FALSE
)
```

- How would you write about the results of this test?

