---
title: "Problem set 1"
editor: source
published-title: "Due date"
date: 2026-04-09
date-modified: last-modified
categories: [problem set]
---

# Purpose

Problem sets are designed for you to test your own conceptual understanding of the material. They provide a skeleton of relevant information from lecture and workshop for you to use as a reference when you are completing homeworks and exams, which are more formal presentations of your understanding.  

# Set up

You may turn in your problem set in .pdf or .docx format. This can be generated using code in a .qmd or .Rmd file, or a regular Word or Google Docs document.    

A **complete** problem set will include:  

- your name and the date
- the title ("Problem set 1")
- answers to each problem component in the listed order (e.g. 1a, 1b, 1c, and so on)
- all answers _visible_ with no missing components

Additionally, a **complete** problem set _must_ be accompanied by a Generative AI statement of use (linked in the portal on Canvas).  

Any problem set missing these components will be marked **incomplete**. See the syllabus for information on how missing problem sets) factor into your course grade.

# Answer keys

_Answers will not be posted._ To compare your problem set against a paper copy of the key, you may:  

- stop by drop-in hours with anyone on the instructional team (Allison, Matt, Abigail, or An)
- make a one-on-one meeting with An (via the meeting link on Canvas) or Allison based on our availability

```{r}
#| echo: false
#| message: false

library(tidyverse)
library(lterdatasampler)
```

# Problems

## 1. Categorizing variables

For each of the following variables, name i) what kind of variable it is and ii) why.  

a. **Date** for a study takes place over 3 days (11 November, 15 December, 13 January) with one observation of frog count per day

b. **Time** for a study that takes place over 24 hours with one measurement of water temperature every 15 minutes (e.g. 04:15, 06:30, 18:45)

c. **Scale of 1-5** in whole numbers where 1 is strongly disagree, 3 is neither agree nor disagree, and 5 is strongly agree

d. **Count** of whale sharks per day along migratory route in Baja California, Mexico (e.g. 4 whale sharks per day, 10 whale sharks per day, 0 whale sharks per day)

e. **Category** of elevation gain on a hike: high, medium, or no elevation gain

f. **Year** for a study that takes place over 32 years, where researchers estimate giant kelp biomass once a year (e.g. 1993, 2001, 2022)

g. **Height** of an oak tree, measured in meters to the nearest whole meter (e.g. 1 m, 4 m, 25 m)

## 2. Central tendency and spread

a. What units would you use for:
  - Mean
  - Median
  - Mode
  - Standard deviation
  - Variance
  - Range
  
b. What functions could you use to calculate:
  - Mean
  - Median
  - Mode
  - Standard deviation
  - Variance
  - Range
  
## 3. Data visualization

a. What is on the x-axis of a histogram? What is on the y-axis of a histogram?

b. What is on the x-axis of a timeseries? What is on the y-axis of a timeseries?

c. Label each component of this boxplot with what it represents mathematically.  

```{r}
#| echo: false

and_vertebrates |> 
  filter(species %in% c("Coastal giant salamander", "Cascade torrent salamander"),
         unittype == "SC") |> 
  ggplot(mapping = aes(x = species,
                       y = length_1_mm)) +
  geom_boxplot() +
  labs(x = "Species",
       y = "Length (mm)") +
  theme_bw()

```

d. If you wanted to compare median vegetation height (cm) across 1) burned grasslands, 2) unburned grasslands, and 3) shrublands, what kind of figure would you create and why?

e. If you wanted to understand the relationship between soil salinity (measured in mS/cm) and nitrogen concentration (measured in parts per million), what kind of figure would you create and why?

f. If you wanted to see how bird species richness (i.e. species number) changed through time, what kind of figure would you create and why?

## 4. Data usage

This is a data set at the Quebrada Sonadora site in the Luqillo Mountains, Puerto Rico measured by researchers at the Luqillo Experimental Forest. The researchers measured stream chemistry parameters every week.


```{r}
#| echo: false

set.seed(666)
luq_streamchem |> 
  select(sample_date, temp, na, no3_n, p_h, ca, k, mg) |> 
  mutate(year = year(sample_date)) |> 
  relocate(year, .after = sample_date) |> 
  sample_n(size = 10) 
``` 

a. What does each column in this data set represent? What does each row represent? Be specific about _time period_ of the observation.

b. There's no metadata (information about the data) available to you, but what is your best guess for what the values in the `na`, `no3_n`, `p_h`, `ca`, `k`, and `mg` columns represent? Be specific regarding the compounds or elements that the researchers may have measured.

c. If you wanted to create a histogram of stream temperature, which column(s) would you use? Which geometry?

d. If you wanted to create a timeseries of calcium concentration through time, which column(s) would you use? Which geometry (name the specific function)?

e. If you wanted to explore the potential relationship between potassium and pH, which column(s) would you use? Which geometry (name the specific function)?

f. If you wanted to create a boxplot comparing magnesium concentration across years, which columns would you use? Which geometry (name the specific function)?

## 5. Metadata

The dataset in the above section on "Data usage" is from the `lterdatasampler` package, but the original dataset is from here:

> McDowell, W. and International Institute of Tropical Forestry(IITF), USDA Forest Service.. 2024. Chemistry of stream water from the Luquillo Mountains ver 4923064. Environmental Data Initiative. https://doi.org/10.6073/pasta/f31349bebdc304f758718f4798d25458 (Accessed 2026-03-26).

All datasets are associated with "metadata", or information about the data.  

Before you start this problem, read [this description of metadata](https://www.ncei.noaa.gov/resources/metadata) to understand what it is.  

Then, click through to the "Introduction to Metadata" page (under "Learn") and read the questions that you should be able to answer with metadata.  

For the stream water chemistry data, answer the following questions based on:

- the data's [metadata](https://doi.org/10.6073/pasta/f31349bebdc304f758718f4798d25458) and 
- the abstract and introduction of the [paper](/data/effects-of-hurricane-disturbance-on-stream-water-concentrations-and-fluxes-in-eight-tropical-forest-watersheds-of-the-luquillo-experimental-forest-puerto-rico.pdf) associated with the dataset

a. Who collected the data?

b. What are the data about?

c. What is the purpose or goal of collecting this data?

d. Where are the data located?

e. When were the data processed?

f. How do I access the data?

