---
title: "Coding workshop: Week 4"
subtitle: "ggplot customization"
categories: [tidyverse, read_csv, ggplot, theme, facet_wrap, geom_point, geom_line, group_by, summarize, pipe operators, '|>', filter, mutate]
format:
  html:
    toc: true
    toc-depth: 8
---

[Workshop date: Friday 30 January]{style="color: #79ACBD; font-size: 24px;"}

## 1. Summary

### Packages
- `tidyverse`  

### Operations

#### Review
- read in data using `read_csv()`  
- visualize data using `ggplot()`  
- modify theme elements using `theme()`  
- use built-in themes using `theme_()`  
- modify colors using `scale_color_()` and `scale_fill_()` functions  
- create multi-panel plots using `facet_wrap()`  
- make line plots using `geom_point()` and `geom_line()`  
- group data using `group_by()`  
- summarize data using `summarize()`  
- chain functions together using ` |> `  
- filtering observations using `filter()`  
- manipulate columns using `mutate()` and `case_when()`  

### Data sources

The fish migration data (`salmon.csv`) is from the [Columbia River DART (Data Access in Real Time)](https://www.cbr.washington.edu/dart/overview) on fish migration through the Columbia River Basin in 2023.  

The Dungeons and Dragons data (`dnd.csv`) is from [insert data source here]().  

The palm tree data (`palmtrees.csv`) is from [insert data source here]().  

The Pokémon data (`pokemon.csv`) is from [insert data source here]().  

The shark incident data (`sharks.csv`) is from [Riley et al.](https://doi.org/10.1038/s41597-022-01453-9)  

The tornado data (`tornados.csv`) is from the [NOAA National Weather Service Storm Prediction Center](https://www.spc.noaa.gov/wcm/#data).  

The data from Mt. Vesuvius (`vesuvius.csv`) is from [insert data source here]().  

## 2. Today's topic

We're going to be exploring different ways to customize plots in `ggplot2` (remember that `ggplot2` is a package within the `tidyverse`, which you already have installed).  

Each group will be responsible for creating a figure manipulating your assigned plot component. You will spend some time exploring each part and creating a figure together. After you are done, you'll compile slides to teach the class  
a) what your assigned plot component is (what it does) and  
b) how to manipulate it (in code).  

**Your challenge is to make the ugliest plot you can!** Change the colors, line types, line widths, etc. - whatever your heart desires to make a fundamentally ugly plot.  

### Before you start coding

#### 1. Look up the `theme()` function.

In the Console, type a question mark, then your function name. For example:

```{r looking-up-function}
#| eval: false # this makes sure that the code shows up in the rendered document but isn't actually run

# copy paste this into the console
?theme

# if you are working on scale_color or scale_fill, type ?scale_color or ?scale_fill into the console and pick one of the options in the pop-up menu
?scale_color
?scale_fill
```

Read about your function. Make a list of the arguments that are relevant to you given your plot element.

#### 2. Read about the `ggplot2` manual section about your plot element.  

If your element is strip, plot, panel, legend, or axis, read the section in ["Modifing theme components"](https://ggplot2-book.org/themes.html#modifying-theme-components) that is relevant to you.  

If your element is `scale_color` or `scale_fill`, read the ["Manual scales" section](https://ggplot2-book.org/scales-other.html#sec-scale-manual).  

If your topic is built-in themes combined with custom theme elements, read the ["Complete themes" section](https://ggplot2-book.org/themes.html#sec-themes).  

#### 3. See your elements in action.

If your element is strip, plot, panel, legend, axis, or if you are combining built-in themes with theme elements, see this tutorial from [ggplot2tor](https://ggplot2tor.com/theme/).  

If your element is `scale_color` or `scale_fill`, see this tutorial from the [R Graph Gallery about dealing with colors](https://r-graph-gallery.com/ggplot2-color.html).  

## 3. Code

### 1. Set up

#### Packages and data

```{r read-in-packages-and-data-background}
#| echo: false
#| message: false

# packages
library(tidyverse)
library(here)

# data
# salmon data
salmon <- read_csv(here("workshop", "data", "adultdaily_1745380588_196.csv"))

# tornado data
tornados <- read_csv(here("workshop", "data", "tornados.csv"))

# shark data
sharks <- read_csv(here("workshop", "data", "sharks.csv"))
```

You will need the `tidyverse`. Insert a code chunk to read that in. Name it `packages`.

```{r}
#| eval: false

library(tidyverse)
```

You will also need to read in your data. Insert a code chunk to read that in. Name it `data`.

:::{.callout-note title="Only read in the data you are working with!"}
You do not need to read in all the datasets. Choose the code for the one you are working with.
:::

```{r read-in-packages-and-data-for-website}
#| eval: false

# salmon data
salmon <- read_csv("salmon.csv")

# dungeons and dragons data
dnd <- read_csv("dnd.csv")

# palm tree data
palmtrees <- read_csv("palmtrees.csv")

# pokemon data
pokemon <- read_csv("pokemon.csv")

# shark data
sharks <- read_csv("sharks.csv")

# tornado data
tornados <- read_csv("tornados.csv")

# mt. vesuvius data
vesuvius <- read_csv("vesuvius.csv")
```

### 2. Cleaning 

For some datasets, you will need to do some cleaning and summarizing.  

If you are working with the `salmon`, `tornados`, or `sharks` data frames, you will need to include the following cleaning/summarizing lines in your .qmd file.  

Insert a new code chunk. Name it `cleaning and wrangling`. Copy-paste the code into your .qmd and run it.

#### Salmon

```{r salmon-data-and-wrangling}
# create new clean object from salmon
salmon_clean <- salmon |> 
  # making sure the date is read as a date
  mutate(Date = mdy(Date)) |> 
  # selecting date and 3 salmonid species
  select(Date, Chin, Stlhd, Coho) |> 
  # making the data frame longer
  pivot_longer(cols = Chin:Coho,
               names_to = "species",
               values_to = "daily_count") |> 
  # mutating species column to display species names in full
  mutate(species = case_when(
    species == "Chin" ~ "Chinook",
    species == "Stlhd" ~ "Steelhead",
    TRUE ~ species
  )) |> 
  # filter to only include dates after December 31st 2023
  filter(Date > as_date("2023-12-31")) |> 
  # take out any missing values
  drop_na(daily_count)
```

#### Tornados

```{r tornados-cleaning-and-wrangling}
# create a new object called tornados_states from tornados
tornados_states <- tornados |> 
  # group by year
  group_by(yr, st) |>
  # calculate total property loss in dollars, sum number of tornados, calculate total property loss in billions of dollars
  summarize(total_property_loss = sum(loss, na.rm = TRUE),
            number_tornados = length(yr),
            total_property_loss_bil = total_property_loss/1000000000) |> 
  # ungroup the data frame (useful if you're going to do any further summarizing steps)
  ungroup() |> 
  # filter state to only include OK, TX, LA, KS
  filter(st %in% c("OK", "TX", "LA", "KS"))
```

#### Sharks

```{r sharks-cleaning-and-wrangling}
# create new clean object from sharks
sharks_clean <- sharks |> 
  # drop any observations where shark common name is missing
  drop_na(shark_common_name) |> 
  # group by shark species and state
  group_by(shark_common_name, state) |> 
  # count observations (easy way to get total number of observations from groups)
  count() |> 
  # ungroup the data frame (good to do if you're doing any cleaning steps after this)
  ungroup() |> 
  # filter to only include New South Wales, Queensland, and Western Australia
  filter(state %in% c("NSW", "QLD", "WA")) |> 
  # mutate state column to show full names for each state
  mutate(state = case_when(
    state == "NSW" ~ "New South Wales",
    state == "QLD" ~ "Queensland",
    state == "WA" ~ "Western Australia"
  ))
```

### 3. Plot customization

Find the plot that accompanies your dataset.  

Create a new code chunk in your .qmd file called `plot customization`.  

Copy paste the code for your plot into your .qmd file to start working with it.

#### a. Salmon

This plot depicts daily counts of salmon through Bonneville Dam in Columbia River Basin, Oregon in 2024. 

```{r salmon-timeseries}
               # base layer: ggplot
salmon_plot <- ggplot(data = salmon_clean,
                      # aesthetics: x-axis, y-axis, and color
                      aes(x = Date,
                          y = daily_count,
                          color = species)) +
  # first layer: points
  geom_point() +
  # second layer: line
  geom_line() +
  # labels
  labs(x = "Date",
       y = "Daily fish count")

# display the plot
salmon_plot
```

#### b. Tornados

Total property loss (in dollars) due to tornados in US from 1950-2022

```{r tornado-property-loss}
tornados_by_state_plot <- ggplot(data = tornados_states,
                                 aes(x = number_tornados,
                                     y = total_property_loss_bil,
                                     color = st)) +
  geom_point()


tornados_by_state_plot
```

#### c. Total number of shark incidents in New South Wales and Queensland from 1791-2022

```{r shark-incident-plot}
              # base layer: ggplot
shark_plot <- ggplot(data = sharks_clean,
                     # aesthetics: x-axis, y-axis
                     aes(x = n,
                         y = shark_common_name)) +
  # first layer: columns to represent counts
  geom_col() +
  # faceting by state
  facet_wrap(~ state) +
  # labels
  labs(x = "Year",
       y = "Total number of shark incidents")

# display the plot
shark_plot
```

#### 3. Plot components

As everyone is going through their plot components, take notes in each section.

###### a. `strip` in `theme()`

Demonstrate how to:  
- change the background  
- change the placement  
- change the text size and font  

**Note:** you may want to use the `shark_plot` for this `theme` element.

```{r strip-elements-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r strip-elements-group}
# insert code here for your group plot
```

###### b. `plot` in `theme()`

Demonstrate how to:  
- change the plot margin  
- change the plot background  
- change the plot title, subtitle, and caption text position and color  

**Note:** you will have to add a title, subtitle, and caption to the plot you choose to manipulate.  

Code for your own independent exploration:

```{r plot-elements-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r plot-elements-group}
# insert code here for your group plot
```

###### c. `panel` in `theme()`

Demonstrate how to:  
- change the panel border  
- change the panel major grid lines (vertically and horizontally, in separate arguments)  
- change the panel minor grid lines (vertically and horizontally, in separate arguments)  
- change the panel background  

Code for your own independent exploration:

```{r panel-elements-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r panel-elements-group}
# insert code here for your group plot
```

###### d. `legend` in `theme()`

Demonstrate how to:  
- change the legend frame  
- change the legend key size  
- change the legend text size  
- change the legend position  
- change the legend row numbers  

**Note:** you may want to use the `salmon_plot` for this `theme` element.  

Code for your own independent exploration:

```{r legend-elements-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r legend-elements-group}
# insert code here for your group plot
```

###### e. `axis` in `theme()`

Demonstrate how to:  
- change the axis text color and font
- change the axis tick length (major and minor ticks)  
- change the axis line colors and line types  

Code for your own independent exploration:

```{r axis-elements-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r axis-elements-group}
# insert code here for your group plot
```

###### f. `scale_color` or `scale_fill` functions

**Note: use the shark plot for scale_fill and any other plot for scale_color**

Demonstrate how to:  
- use a color palette package  
- apply it to a color scale  

Code for your own independent exploration:

```{r color-or-fill-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r color-or-fill-group}
# insert code here for your group plot
```

###### g. built in themes (`theme_`) with your own customization using `theme()`

Demonstrate how to:  

- use a built in theme and  
- change additional components using the `theme()` elements of your choice

Code for your own independent exploration:

```{r theme-customization-individual}
# insert code here for your individual plot
```

Code for the plot your group made:

```{r theme-customization-group}
# insert code here for your group plot
```

## 3. Extra stuff

### Thursday 4 PM plot

```{r thursday-plot}
#| fig-width: 12
#| fig-height: 7
#| code-fold: true

ggplot(data = sharks_clean,
                     # aesthetics: x-axis, y-axis
                     aes(x = n,
                         y = shark_common_name,
                         # fill columns by shark common name
                         fill = shark_common_name)) +
  # first layer: columns to represent counts
  geom_col() +
  # faceting by state, putting the strip on the bottom of the plot
  facet_wrap(~ state,
             strip.position = "bottom") +
  # labels
  labs(x = "Total number of shark incidents",
       y = "Species") +
  # manually manipulate fill colors
  scale_fill_manual(values = c("bull shark" = "pink",
                               "whaler shark" = "chartreuse")) +
  # built in theme
  theme_dark() +
  # theme elements
  theme(
    # strip elements
    strip.placement = "outside",
    strip.background = element_rect(color = "red", linetype = 4, linewidth = 7),
    strip.text = element_text(color = "hotpink"),
    strip.text.x.top = element_text(size = 20, face = "italic"),
    # plot elements
    plot.background = element_rect(fill = "turquoise", color = "#EC4899", linewidth = 4, linetype = "dashed"),
    # panel elements
    panel.grid.major.x = element_line(color = "hotpink", linewidth = 3),
    panel.grid.major.y = element_line(color = "dodgerblue", linewidth = 1),
    panel.background = element_rect(fill = "lavender"),
    # legend elements
    legend.background = element_rect(fill = "chartreuse"),
    legend.key = element_rect(color = "orange"),
    legend.text = element_text(size = 30),
    # axis elements
    axis.line.x = element_line(color = "yellowgreen", linewidth = 4),
    axis.line.y = element_line(color = "deeppink1"),
    axis.text.x.bottom = element_text(color = "skyblue"),
    axis.text.y.left = element_text(color = "chocolate")
  )
```

### Thursday 5 PM plot

```{r}
#| fig-width: 12
#| fig-height: 7
#| code-fold: true

ggplot(data = salmon_clean,
                      # aesthetics: x-axis, y-axis, and color
                      aes(x = Date,
                          y = daily_count,
                          color = species)) +
  # first layer: points
  geom_point() +
  # second layer: line
  geom_line(size = 2, 
            linetype = "dashed") +
  # labels
  labs(x = "Date",
       y = "Daily fish count",
       title = "Daily fish count per day",
       subtitle = "group 2",
       caption = "thursday 5 pm") +
  # custom color scale
  scale_color_manual(values = c("Chinook" = "hotpink",
                                "Coho" = "chartreuse",
                                "Sockeye" = "pink")) +
  # adding this facet 
  facet_wrap(~species) +
  # built in theme
  theme_dark() +
  theme(
    # strip elements
    strip.text = element_text(family = "Times New Roman"),
    strip.background = element_rect(fill = "red"),
    # plot elements
    plot.title = element_text(size = 30, face = "bold", color = "orange"),
    plot.subtitle = element_text(color = "yellow"),
    plot.margin = margin(t = 100, r = 0.6, b = 20, l = 0.9),
    plot.background = element_rect(fill = "lightgreen"),
    # panel elements
    panel.grid.minor.x = element_line(color = "orange", linewidth = 1),
    panel.grid.major.y = element_line(color = "darkgreen", linewidth = 5),
    panel.background = element_rect(fill = "black", color = NA),
    # legend elements
    legend.direction = "horizontal",
    legend.key.size = unit(3, "line"),
    legend.background = element_rect(fill = "lemonchiffon", color = "grey50", linewidth = 1),
    # axis elements
    axis.text.y.left = element_text(size = 15, color = "deeppink")
  )
```

### Friday 9 AM plot

```{r}
#| fig-width: 12
#| fig-height: 7
#| code-fold: true

ggplot(data = sharks_clean,
                     # aesthetics: x-axis, y-axis
                     aes(x = n,
                         y = shark_common_name)) +
  # first layer: columns to represent counts
  geom_col() +
  # faceting by state
  facet_wrap(~ state,
             strip.position = "bottom") +
  # labels
  labs(x = "Total number of shark incidents",
       y = "Shark species",
       title = "Fishy Salmon Run",
       subtitle = "fishfishyfishyfishyfishy") +
  # built in theme
  theme_dark() +
  theme(
    # strip elements
    strip.text = element_text(color = "brown",
                              family = "Times New Roman",
                              size = 50,
                              face = "bold"),
    strip.background = element_rect(fill = "cornflowerblue",
                                    color = "pink2",
                                    size = 5),
    # plot elements
    plot.background = element_rect(fill = "#EC4899",
                                   color = NA),
    plot.title = element_text(angle = -20, size = 40, color = "blue"),
    plot.subtitle = element_text(angle = 20, color = "gold"),
    plot.margin = unit(rep(2, 4), "cm"),
    # panel elements
    panel.grid.major = element_line(size = 8),
    panel.background = element_rect(fill = "#EC4899"),
    panel.grid.minor = element_line(color = "#32CD32"),
    # legend elements
    legend.background = element_rect(fill = "yellow"),
    legend.key.size = unit(3, "cm"),
    legend.position.inside = c(0.6, 0.5)
  )
```

### Friday 10 AM plot

```{r}
#| fig-width: 12
#| fig-height: 7
#| code-fold: true

ggplot(data = salmon_clean,
                      # aesthetics: x-axis, y-axis, and color
                      aes(x = Date,
                          y = daily_count,
                          color = species)) +
  # first layer: points
  geom_point() +
  # second layer: line
  geom_line(size = 2, 
            linetype = "dashed") +
  # labels
  labs(x = "Date",
       y = "Daily fish count",
       title = "Ugly plot",
       caption = "super ugly") +
  # custom color scale
  scale_color_manual(values = c("Chinook" = "hotpink",
                                "Coho" = "chartreuse",
                                "Sockeye" = "pink")) +
  # adding this facet 
  facet_wrap(~species) +
  # built in theme
  theme_dark() +
  theme(
    # strip elements
    strip.text = element_text(family = "Courier New",
                              color = "gold",
                              face = "bold"),
    strip.background = element_rect(color = "brown",
                                    fill = "white"),
    # plot elements
    plot.background = element_rect(fill = "yellow", color = "red", size = 20),
    plot.title = element_text(color = "purple", size = 28, face = "italic", hjust = 0),
    plot.caption = element_text(color = "red", size = 14, face = "bold.italic", hjust = 0.5),
    # panel elements
    panel.border = element_rect(fill = NA, color = "chartreuse", size = 1),
    panel.grid.major.x = element_line(color = "blue3"),
    panel.grid.minor.y = element_line(color = "chartreuse4"),
    panel.background = element_rect(fill = "deeppink"),
    # legend elements
    legend.key.height = unit(3, "cm"),
    legend.key.width = unit(5, "cm"),
    legend.background = element_rect(fill = "green"),
    legend.position = "left",
    legend.box.background = element_rect(fill = "red"),
    # axis elements
    axis.text.y.left = element_text(size = 17, color = "yellow"),
    axis.ticks = element_line(color = "red"),
    axis.line = element_line(color = "purple",
                             size = 12)
  )
```

