---
title: "Coding workshop: Week 5"
subtitle: "Comparisons of more than two groups: ANOVA and Kruskal-Wallis"
date: 2026-05-01
categories: [tidyverse, lterdatasampler, rstatix, aov, summary, TukeyHSD, eta_squared, read_csv, pipe operators, '|>', select, rename, ggplot, geom_histogram, geom_qq, geom_qq_line, facet_wrap, group_by, summarize, fct_relevel, mutate]
format:
  html:
    toc: true
    toc-depth: 8
---

[Workshop date: Friday 1 May]{style="color: #79ACBD; font-size: 24px;"}

## 1. Summary

### Packages

- `tidyverse`
- `lterdatasampler`
- `rstatix`

### Operations

#### New functions

- make sure a variable is a factor using `as_factor()`
- do an ANOVA using `aov()`
- look for more information from model results using `summary()`
- do post-hoc Tukey test using `TukeyHSD()`
- calculate effect size for ANOVA using `rstatix::eta_squared()`

#### Review

- read in data using `read_csv()`
- chain functions together using ` |> `
- modify columns using `mutate()`
- select columns using `select()`
- rename columns using `rename()`
- set factor level order using `fct_relevel()`
- visualize data using `ggplot()`
- create histograms using `geom_histogram()`
- visualize QQ plots using `geom_qq()` and `geom_qq_line()`
- create multi-panel plots using `facet_wrap()`
- group data using `group_by()`
- undo grouping using `ungroup()`
- summarize data using `summarize()`

### Data sources

The Plum Island Ecosystem fiddler crab data is from `lterdatasampler` (data info [here](https://lter.github.io/lterdatasampler/articles/pie_crab_vignette.html)). Crab carapace width is measured in mm.

## 2. Code

### 1. Set up

#### Packages

```{r}
#| label: packages
#| message: false

library(tidyverse)
library(lterdatasampler)
library(rstatix)
```

#### Data 
```{r}
#| label: data-for-website
#| eval: false

# for parametric tests
data(pie_crab)

# for non-parametric tests
ramen_ratings <- read_csv("ramen_ratings.csv")
```

```{r}
#| label: data
#| echo: false
#| message: false

# for parametric tests
data(pie_crab)

# for non-parametric tests
ramen_ratings <- read_csv(here::here("workshop", "data", "ramen_ratings.csv"))
```

### 2. Parametric tests

#### a. Cleaning and wrangling

- clean data frame from `pie_crab`
- filter to only include: Cape Code, Zeke's Island, and Virginia Coastal LTER
- select columns we need: `site`, `name`, and `size`
- rename columns to reflect values in the columns 

```{r}
#| label: crab-cleaning

# start with the pie_crab dataset
pie_crab_clean <- pie_crab |> 
  # filter for Cape Cod, Zeke's Island, Virginia Coastal
  filter(site %in% c("CC", "ZI", "VCR")) |> 
  # select columns of interest
  select(site, name, size) |> 
  # rename columns
  rename(site_code = site, 
         site_name = name) 
```

#### b. Quick summary

- create a new object to store summary
- calculate summary values: mean, variance, and sample size

```{r}
#| label: crab-summary

# creating a new object called pie_crab_summary
pie_crab_summary <- pie_crab_clean |> 
  # group by site
  group_by(site_name) |> 
  # calculate mean, variance, sample size
  summarize(mean = mean(size), 
            var = var(size), 
            n = length(size)) |> 
  # ungroup data frame
  ungroup()

# display pie_crab_summary
pie_crab_summary
```

#### c. Exploring the data

Create a jitter plot with the mean crab size for each site.

```{r}
#| label: exploratory visualization

# base layer: ggplot
ggplot(data = pie_crab_clean,
       # x-axis: site
       # y-axis: crab size
       mapping = aes(x = site_name,
                     y = size)) +
  # first layer: jitter to represent individual crabs
  geom_jitter(height = 0,
              width = 0.2,
              shape = 21) +
  # second layer: points to represent site means
  stat_summary(geom = "point",
               fun = mean,
               color = "red",
               size = 3) +
  # using a different theme from the default
  theme_minimal() +
  # relabelling the x- and y-axes
  labs(x = "Site",
       y = "Crab carapace width (mm)")
```

Is there a difference in mean crab carapace width (mm) between the three sites? What visual components of the plot are you using to make that guess? 

**There seems to be a difference in mean crab carapace widths between sites. Cape Cod and Virginia Coastal Reserve crabs seem to be similar, and Zeke's Island crabs are smaller than the other two.**

#### d. Check 1: normally distributed variable

Create a histogram of crab size. Facet your histogram so that you have 3 panels, with one panel for each site.

```{r}
#| label: crab-histogram

# base layer: ggplot
ggplot(data = pie_crab_clean,
       # x-axis: crab carapace width
       mapping = aes(x = size)) +
  # histogram
  geom_histogram(bins = 9,
                 fill = "tomato4",
                 color = "black") +
  # using a different theme
  theme_minimal() +
  # creating different panels for each site
  facet_wrap(~ site_name) +
  # theme options
  theme(text = element_text(size = 14))
```

Create a QQ plot of crab size. Facet your QQ plot so that you have 3 panels, with one panel for each site.

```{r}
#| label: crab-qq

# base layer: ggplot
ggplot(data = pie_crab_clean,
       # y-axis: crab carapace size
       mapping = aes(sample = size)) +
  # reference line for the QQ
  geom_qq_line(alpha = 0.5) +
  # points representing sample vs theoretical
  geom_qq(color = "tomato4") +
  # different theme
  theme_minimal() +
  # creating different panels for each site
  facet_wrap(~ site_name) +
  # theme options
  theme(text = element_text(size = 14))
```

**Note**: Crab carapace width seems to be normally distributed based on visual checks using histograms and QQ plots.

**Writing**: We evaluated normality of carapace width using histograms and QQ plots.

#### e. Check 2: equal variances

Do a gut check: is the ratio of the largest variance to the smallest variance less than 3?

```{r}
#| label: crab-variance

8.63/4.04
```

**Note**: the variances are _equal enough_ for the purposes of ANOVA based on the rule of thumb: the ratio of the largest to the smallest variance should be less than 3.

#### f. ANOVA

```{r}
#| label: crab-anova

# creating an object called crab_anova
crab_anova <- aov(
  # size as a function of site
  size ~ site_name, 
  # using clean data frame
  data = pie_crab_clean
) 

# object
crab_anova

# gives more information
summary(crab_anova)
```

**Note**: There is a difference in mean crab carapace width between Cape Cod, Zeke's Island, and Virginia Coastal Reserve.

#### g. Post-hoc: Tukey HSD

```{r}
#| label: crab-tukey

# post-hoc for ANOVA
TukeyHSD(crab_anova)
```

Which pairwise comparisons are actually different? Which ones are not different?  

**Zeke's Island NERR and Cape Cod crabs are different, and Zeke's Island NERR and Virginia Coastal Reserve LTER crabs are different. Virginia Coastal Reserve LTER and Cape Cod crabs are not different.**

#### h. effect size

Using `eta_squared()` from `rstatix`

```{r}
#| label: crab-effectsize

rstatix::eta_squared(crab_anova)
```

What is the magnitude of the differences between sites in crab size?  

**There is a large difference in mean crab carapace width between sites.**  

Another way of saying this: there is a **large** effect of site on crab carapace width. There is a **large** difference between sites in crab carapace width.

#### i. Putting everything together

**means**

CC: 16.8 mm, VCR: 16.3 mm, ZI: 12.1 mm

**sample size:**

CC: 27; VCR: 30, ZI: 35

**test type**

one-way ANOVA

**test distribution**

F distribution

**distribution parameters**

numerator df: 2; denominator df: 89

**test statistic**

39.6

**p-value**

5.1e-13 (expressed as p < 0.001)

**significance level**

$\alpha$ = 0.05

**post-hoc results**

VCR and CC are not different, but Zeke's Island and Cape Cod are different and Zeke's Island and VCR are different.

**effect size**

$\eta^2$ = 0.47

**Writing**

We found a large difference ($\eta^2$ = 0.47) in mean crab carapace widths between sites (one-way ANOVA, F(2, 89) = 39.6, p < 0.001, $\alpha$ = 0.05). The smallest crabs were from Zeke's Island NERR, which had an average width of 12.1 mm (n = 35). Zeke's Island crabs were 4.3 mm smaller than crabs from Virginia Coastal Reserve (VCR n = 30, 95% CI: [3.3, 6.2] mm difference) and 4.8 mm smaller than crabs from Cape Cod (CC n = 27, 95% CI: [2.9, 5.7] mm difference).

### 3. Non-parametric tests

#### a. Clean and wrangle the data

```{r}
#| label: ramen-wrangle

# creating new object from ramen_ratings
ramen_ratings_clean <- ramen_ratings |> 
  # filtering to only include Maruchan brand
  filter(brand == "Maruchan") |> 
  # changing level orders in style
  mutate(style = fct_relevel(style,
                             "Cup", "Pack", "Tray", "Bowl"))

# display structure of data frame
str(ramen_ratings_clean)
```

#### b. Quick summary

```{r}
#| label: ramen-summary

# creating new object from ramen_ratings_clean
ramen_ratings_summary <- ramen_ratings_clean |> 
  # grouping by style
  group_by(style) |> 
  # calculating the median and sample size
  summarize(median = median(stars),
            n = length(stars)) |> 
  # ungrouping data frame
  ungroup()

# display the object
ramen_ratings_summary
```


#### c. Make a boxplot to compare star ratings across ramen styles

```{r}
#| label: ramen-exploratory-visualization

# base layer: ggplot
ggplot(data = ramen_ratings_clean,
       # x-axis: style,
       # y-axis: star ratings
       # color geometries by style
       mapping = aes(x = style,
                     y = stars,
                     color = style)) +
  # first layer: boxplot representing medians
  geom_boxplot(outliers = FALSE) +
  # second layer: jitter representing individual ramen types
  geom_jitter(height = 0,
              width = 0.2) +
  # using custom colors
  scale_color_manual(values = c(
    "Cup" = "firebrick4",
    "Pack" = "orange",
    "Tray" = "goldenrod",
    "Bowl" = "darkgreen"
  )) +
  # using a theme that's different from default
  theme_minimal() +
  # taking out the legend
  theme(legend.position = "none")
```

#### d. Do the Kruskal-Wallis test

```{r}
#| label: ramen-kw

kruskal.test(
  # formula: star ratings as a function of style
  stars ~ style,
  # data: clean ramen ratings data frame
  data = ramen_ratings_clean
)
```

Is there a difference in ratings between ramen styles?  

**Ramen styles differ in median ratings.**

#### e. Do a Dunn's post-hoc test

Using `dunn_test()` from `rstatix`

```{r}
#| label: ramen-post-hoc

dunn_test(
  # formula: star ratings as a function of style
  stars ~ style,
  # data: clean ramen ratings data frame
  data = ramen_ratings_clean
)
```

Which pairwise comparisons of ramen styles are different from each other?  

**Cups and bowls are different, packs and bowls are different, and tray and bowls are different.**

#### f. Calculate an effect size

Using `kruskal_effsize()` from `rstatix`

```{r}
#| label: ramen-effectsize

kruskal_effsize(
  # formula: star ratings as a function of ramen style
  stars ~ style, 
  # data frame
  data = ramen_ratings_clean 
)
```

What is the magnitude of the effect of ramen style on ratings?  

**There is a moderate difference in median star ratings between ramen styles.**

#### g. Putting everything together

**medians**

cup: 3.5 stars, pack and tray: 3.75 stars, bowl: 4.25 stars

**sample sizes**

cup: 26, pack: 30, tray: 17, bowl: 33

**test**

Kruskal-Wallis

**test "distribution"**

$\chi^2$

**distribution parameters**

degrees of freedom: 3

**test statistic**

15.7

**p-value**

0.001

**Dunn's test**

cups and bowls are different, packs and bowls are different, and tray and bowls are different

**effect size**

$\eta^2$ = 0.12

**Writing**

We found a moderate difference ($\eta^2$ = 0.12) between ramen styles in median star rating (Kruskal-Wallis, $\chi^2$ = 15.7, p = 0.001, $\alpha$ = 0.05). Bowl style ramen tended to be rated highest, with a median star rating of 4.25 stars (n = 33). Bowls tended to be rated higher than packs (pack median: 3.75 stars, n = 30, Holm adjusted p-value = 0.04), trays (tray median: 3.75 stars, n = 17, Holm adjusted p-value = 0.04), and cups (cup median: 3.5 stars, n = 26, Holm adjusted p-value = 0.001).



### 4. Captions

#### Figure 1.

**Figure 1. Crab carapace width differs across sites.** Small black points represent measurements of crab carapace widths at three sites: Zeke's Island National Estuarine Research Reserve (NERR), Virginia Coastal Reserve Long-Term Ecological Research (LTER) site, and Cape Cod. Large red points represent mean crab carapace width. Data originally from: Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb. Accessed via `lterdatasampler` package.

#### Figure 2. 

**Figure 2. Histograms of crab carapace width.** Columns represent counts of individual crabs. Distributions of crab carapace width are represented in three panels: Zeke's Island NERR (left), Virginia Coastal Reserve LTER (center), and Cape Cod (right). Data originally from: Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb. Accessed via `lterdatasampler` package.  

Alternative option if you have a figure made using the same dataset as a previous figure:  

**Figure 2. Histograms of crab carapace widths.** Bars represent counts of crab sizes. Distributions are shown for three sites: Zeke's Island NERR (left), Virginia Coastal Reserve (center), and Cape Cod (right). Data source and access as in Figure 1. 

#### Figure 3.

**Figure 3. Quantile-Quantile (QQ) plots of crab carapace width.** Points represent individual observations of crab carapace width. Lines represent reference line relating theoretical quantiles (x-axis) and sample quantiles (y-axis). Panels represent different sites: Zeke's Island NERR (left), Virginia Coastal Reserve LTER (center), and Cape Cod (right). Data originally from: Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb. Accessed via `lterdatasampler` package.  

Alternative option if you have a figure made using the same dataset as a previous figure: 

**Figure 3. Quantile-quantile (QQ) plots of crab carapace widths.** Points represent individual crab carapace width measurements. Lines represent reference quantiles of sample (y-axis) and standard normal theoretical (x-axis) distributions. Panels, data source, and access as in Figure 2.

#### Figure 4. 

**Figure 4. Median star ratings for Maruchan Ramen styles.** Points represent individual ratings for ramen types in each style (red: bowl, orange: pack, yellow: tray, green: cup). Boxplots represent median star ratings in each style.

**END OF WORKSHOP 5**

## Extra stuff

### More applications of `stat_summary()`

`stat_summary()` in the `ggplot2` package is a useful function to do quick summary visualizations within a plot.  

In class, we used `stat_summary()` to plot mean carapace widths at each site in red (see the "Exploring the data" section above).  

You can also use `stat_summary()` to visualize means and bars (representing standard deviation, standard error, confidence intervals, etc.).  

For example, here is a plot using `stat_summary()` to visualize means and standard errors:

```{r}
#| label: mean and SD

# base layer: ggplot
ggplot(data = pie_crab_clean, # use the clean data set
       mapping = aes(x = site_name, # x-axis
                     y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "pointrange", # geometry being plotted
               fun.data = mean_sdl, # function to calculate mean and sd (note new argument name)
               color = "red", # color to make it easier to see
               size = 1) + # larger point size
  theme_minimal() + # cleaner plot theme
  labs(x = "Site name",
       y = "Carapace width (mm)",
       caption = "Mean \U00B1 standard deviation")

```

And standard error:

```{r}
#| label: mean and SE

# base layer: ggplot
ggplot(data = pie_crab_clean, # use the clean data set
       mapping = aes(x = site_name, # x-axis
                     y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "pointrange", # geometry being plotted
               fun.data = mean_se, # function to calculate mean and SE (note new argument name)
               color = "red", # color to make it easier to see
               size = 1) + # larger point size
  theme_minimal() + # cleaner plot theme
  labs(x = "Site name",
       y = "Carapace width (mm)",
       caption = "Mean \U00B1 standard error")

```

And a 95% confidence interval:

```{r}
#| label: mean and confidence interval

# base layer: ggplot
ggplot(data = pie_crab_clean, # use the clean data set
       mapping = aes(x = site_name, # x-axis
                     y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "pointrange", # geometry being plotted
               fun.data = mean_cl_normal, # function to calculate range for confidence interval (note new argument name)
               color = "red", # color to make it easier to see
               size = 1) + # larger point size
  theme_minimal() + # cleaner plot theme
  labs(x = "Site name",
       y = "Carapace width (mm)",
       caption = "Means with 95% confidence interval")

```









