---
title: "Coding workshop: Week 3"
subtitle: "basic t-tests and their assumptions"
date: 2026-04-17
categories: [tidyverse, lterdatasampler, data, geom_qq, geom_qq_line, facet_wrap, var.test, t.test, rownames_to_column, geom_pointrange, pipe operators, '|>', mutate, case_when, gglpot, geom_boxplot, geom_jitter, geom_histogram, group_by, ungroup, summarize]
---

[Workshop date: Friday 17 April]{style="color: #79ACBD; font-size: 24px;"}

## 1. Summary

### Packages

- `tidyverse`
- `lterdatasampler`

### Operations

#### New functions

- display data from package using `data()`
- visualize QQ plots using `geom_qq()` and `geom_qq_line()`
- create multi-panel plots using `facet_wrap()`
- compare group variances using `var.test()`
- do t-tests using `t.test()`
- make rownames into a separate column using `rownames_to_column()`
- use `geom_pointrange()` to show means and 95% CI

#### Review

- chain functions together using ` |> `
- filtering observations using `filter()`
- manipulate columns using `mutate()` and `case_when()`
- visualize data using `ggplot()`
- create boxplots using `geom_boxplot()` and show observation values using `geom_jitter()`
- create histograms using `geom_histogram()`
- group data using `group_by()`
- undo grouping using `ungroup()`
- summarize data using `summarize()`

### General Quarto formatting tips

You can control the appearance of text, links, images, etc. using this [guide](https://quarto.org/docs/authoring/markdown-basics.html).

### Data source

The data on sugar maples is from the `lterdatasampler` package. The package developers (alumni of the Bren Masters of Environmental Data Science program!) curated a bunch of datasets from the LTER network into this package for teaching and learning. Read about the package [here](https://lter.github.io/lterdatasampler/index.html).  

The source of the data is Hubbard Brook Experimental Forest. Read more about the data [here](https://lter.github.io/lterdatasampler/articles/hbr_maples_vignette.html).

## 2. Code

**Remember to set up an Rproject before starting!**

### 1. Set up

Insert a code chunk below to read in your packages. Name the code chunk `packages`.

```{r packages}
#| message: false # remember to include this in your code chunk whenever you want to make sure that the code shows up in your rendered document without messages

library(tidyverse)
library(lterdatasampler)
```

Because we are using data from the package `lterdatasampler`, we don't need to use `read_csv()`.  

Instead, we can use `data()` to make the data frame show up in the environment.  

Insert a code chunk below to display `hbr_maples` in the environment using `data("hbr_maples")`. Name the code chunk `data`.

```{r data}
#| message: false

data("hbr_maples")
```

### 2. Cleaning and wrangling

Insert a code chunk to:  

1. create a new object from `hbr_maples` called `maples_2003`  
2. filter observations to only include the year 2003  
3. mutate the `watershed` column so that `W1` is filled in as `Calcium-treated`  

Name the code chunk `data cleaning`.

```{r data-cleaning}
# start with hbr_maples data frame
maples_2003 <- hbr_maples |> 
   # filter to only include observations from 2003
  filter(year == "2003") |>
  # rename watersheds
  mutate(watershed = case_when( 
    watershed == "Reference" ~ "Reference",
    watershed == "W1" ~ "Calcium-treated"
  ))
```

### 3. Exploratory data visualization

Insert a code chunk to make a boxplot + jitter plot comparing stem lengths between watersheds. Remember to:  

1. color by watershed  
2. control the jitter so that the points don't move up and down the y-axis  

Name the code chunk `boxplot-with-jitter`.

```{r}
#| label: boxplot-with-jitter

# base layer: ggplot
ggplot(data = maples_2003, # starting data frame
       mapping = aes(x = watershed, # x-axis
                     y = stem_length, # y-axis
                     color = watershed)) + # coloring by watershed
  # first layer: boxplot
  geom_boxplot() +
  # second layer: jitter plot
  geom_jitter(height = 0, # making sure points don't move along y-axis
              width = 0.2) # narrowing width of jitter
```

### 4. Checks for t-test assumptions

#### a. Normally distributed response variable

##### Histogram

Insert a code chunk to create a histogram. Name the code chunk `histogram`.  

Use `facet_wrap()` to create separate panels for each watershed. 

```{r}
#| label: histogram

# starting data frame
ggplot(data = maples_2003, 
       # x-axis (no y-axis for histogram)
       mapping = aes(x = stem_length)) + 
  # number of bins from Rice Rule
  geom_histogram(bins = 6) + 
  # creating two panels to show watersheds separately
  facet_wrap(~watershed) 
```

##### QQ plot

Insert a code chunk to create a QQ plot. Name the code chunk `qq-plot`.  

Use `facet_wrap()` to create separate panels for each watershed.

```{r}
#| label: qq-plot

# base layer: ggplot call
ggplot(data = maples_2003,
       # y-axis for QQ plot (no x-axis for QQ plot)
       mapping = aes(sample = stem_length)) + 
  # first layer: QQ reference line in blue
  geom_qq_line(color = "blue") +
  # second layer: QQ plot
  geom_qq() + 
  # creating "facets"
  facet_wrap(~ watershed) # show watersheds separately
```

**Check in:** using histograms and QQ plots, does stem length seem to be normally distributed?  

**Yes, because the histogram looks symmetrical, and the QQ plot points follow a straight line.**  

#### b. Equal variances

Next, we'll check our variances. 

##### Variance ratio by hand

We can make sure we know where the F test results are coming from by calculating the variance ratios ourselves. 

```{r}
#| label: variance-ratio-calculation

# create new object from maples_2003
stem_length_var <- maples_2003 |>
  # group by watershed
  group_by(watershed) |> 
  # calculate variances
  summarize(variance = var(stem_length)) |> 
  # ungroup data frame
  ungroup()

# calculate variance ratio
# double check against results of var.test()
205.7026/194.3021
```

##### F-test of equal variances

Insert a code chunk to check the variances using `var.test()`. Name the code chunk `F-test`.  

In the function `var.test()`, enter the arguments for:  

1. the formula  
2. the data  

```{r}
#| label: F-test

# doing F test of equal variances
var.test(
  # formula: response variable ~ grouping variable
  stem_length ~ watershed, 
  # data: maples_2003 data frame
  data = maples_2003 
)
```

Remember that this variance test is an F test of equal variances. You are comparing the variance of one group with another.  

To communicate about this, you could write something like:  

Using an F-test of equal variances, we determined that variances were (equal or not equal) (F(num df, denom df) = F ratio, p-value).  

We determined that group variances were (equal or not equal) (F-test of equal variances, F(num df, denom df) = F statistic, p-value).  

Fill in the blank here:  

**We determined that group variances were equal (F-test of equal variances, F(119, 119) = 1.06, p = 0.76, $\alpha = 0.05$).**    

### 5. Doing a t-test

Insert a code chunk to do a t-test. Name the code chunk `t-test`.  

In the function `t.test()`, enter the arguments for:  

1. the formula  
2. the variances in `var.equal = `  
3. and the dataframe in `data = `  

```{r}
#| label: t-test

t.test(
  # formula: response variable ~ grouping variable
  stem_length ~ watershed, 
  # data: maples_2003 data frame
  data = maples_2003, 
  # variances are equal (checked using F-test)
  var.equal = TRUE 
)
```

### 6. Communicating

#### a. visual communication

When doing a t-test, remember that you are comparing _means_. To visualize the data in a way that reflects the values you are comparing (again, you are comparing _means_), you can visualize the _means_ of each watershed with the standard deviation (spread), standard error (variation), or confidence interval (confidence).  

In this example, we will show 95% confidence intervals.  

In this code chunk, we are calculating the means and 95% confidence intervals. Name the code chunk `ci-calculation`.

```{r}
#| label: ci-calculation

# creating new object from maples_2003
maples_ci <- maples_2003 |> 
  # group by watershed
  group_by(watershed) |> 
  # calculations
  summarize(mean = mean(stem_length), # mean
            n = length(stem_length), # number of observations
            df = n - 1, # degrees of freedom
            sd = sd(stem_length), # standard deviation
            se = sd/sqrt(n), # standard error
            tval = qt(p = 0.05/2, df = df, lower.tail = FALSE), # t-value
            margin = tval*se, # margin of error
            ci_lower = mean - margin, # lower bound of the confidence interval
            ci_upper = mean + margin # upper bound of the confidence interval
          ) |> 
  # ungrouping the data frame
  ungroup()
```

Before moving on, look at the `maples_ci` object to make sure you know what it contains.  

Note that this visualization uses two data frames. We use `maples_2003` to show the underlying data using `geom_jitter()`, and `maples_ci` to show the mean and 95% CI.

```{r}
#| label: ci-visualization

# base layer: ggplot with the x- and y-axes
ggplot(data = maples_2003, # using the maples_2003 data frame
       mapping = aes(x = watershed, # x-axis
                     y = stem_length, # y-axis
                     color = watershed)) + # coloring points by watershed
  # first layer: showing the underlying data
  geom_jitter(height = 0, # no jitter in the vertical direction
              width = 0.1, # smaller jitter in the horizontal direction
              alpha = 0.4, # make the points more transparent
              shape = 21) + # make the points open circles
  # second layer: showing the summary (mean and 95% CI)
  # using maples_ci data frame
  geom_pointrange(data = maples_ci, 
                  aes(x = watershed, 
                      y = mean, 
                      ymax = ci_lower, 
                      ymin = ci_upper)) + 
  # relabelling x- and y-axes
  labs(x = "Watershed", 
       y = "Stem length (cm)") +
  # figure customization to change point colors
  # Note: this is optional (but nice to do!)
  scale_color_manual(values = c("Calcium-treated" = "darkorchid3",
                                "Reference" = "tomato3")) + 
  # using a theme
  theme_bw() + 
  # getting rid of the legend
  theme(legend.position = "none") 
```

#### b. Writing

Summarize the results of the t-test in one sentence. Before you do, make sure you know the:  

##### type of test  

Student's t (note: this is because variances are equal)

##### Sample size  

n = 120 for calcium-treated, n = 120 for reference (240 total)  

##### Sample means

87.9 cm for calcium-treated, 81.0 cm for reference

##### significance level ($\alpha$)  

$\alpha$ = 0.05 

##### degrees of freedom  

238 (note: 240 - 2 = 238)  

##### t-value (aka t-statistic)  

3.8

##### p-value  

p < 0.001 (don't need to give exact number for anything below 0.001)  

##### Results summary

We found a significant difference in sugar maple stem lengths between calcium-treated (n = 120, mean = 87.9 cm) and reference (n = 120, mean = 81.0 cm) watersheds (Student's t-test, t(238) = 3.8, p < 0.001, $\alpha$ = 0.05).      

**END OF WORKSHOP 3**

## 3. Extra stuff

### Setting global options for code execution

If you want _any_ messages to be hidden in your rendered Quarto document, then you can adjust the **YAML** at the top of the document. This is the section that has 3 dashes at the beginning and end (where you enter your name, the date, and your execution option).  

The **YAML** controls the aesthetics of the rendered doc (or PDF, or HTML) that is created from your Quarto document. See more details [here](https://rpubs.com/drgregmartin/1266674) and [here from the Quarto user manual](https://quarto-tdg.org/yaml).  

To adjust the YAML for a .qmd file for which you want all messages to be hidden, you can add the following lines to the YAML:

```{r}
#| eval: false

execute:
  message: false
```

So the complete YAML for workshop 3, for example, would look something like this:

```{r}
#| eval: false

---
title: "Week 3 workshop: sugar maples"
author: "An Bui"
format: docx
execute:
  message: false
---
```

:::{.callout-note title="Indentations matter!" collapse="true"}
When adjusting settings in the YAML, you need to be very careful about the indentations. The YAML code as written above will work, but YAML code written without the tab before `message` (example below) will not.  

```{r}
#| eval: false

---
title: "Week 3 workshop: sugar maples"
author: "An Bui"
format: docx
execute:
message: false
---
```

:::
