---
title: "Coding workshop: Week 5"
subtitle: "Analysis of variance, figures and captions"
categories: [tidyverse, lterdatasampler, effectsize, car, shapiro.test, leveneTest, aov, summary, TukeyHSD, eta_squared, read_csv, pipe operators, '|>', select, rename, ggplot, geom_histogram, geom_qq, geom_qq_line, facet_wrap, group_by, summarize, as_factor, mutate]
format:
  html:
    toc: true
    toc-depth: 8
---

[Workshop date: Friday 6 February]{style="color: #79ACBD; font-size: 24px;"}

## 1. Summary

### Packages
- `tidyverse`  
- `lterdatasampler`  
- `rstatix`  
- `car`  

### Operations

#### New functions

- make sure a variable is a factor using `as_factor()`
- do a Shapiro-Wilk test using `shapiro.test()`  
- do a Levene's test using `car::leveneTest()`  
- do an ANOVA using `aov()`  
- look for more information from model results using `summary()`  
- do post-hoc Tukey test using `TukeyHSD()`  
- calculate effect size for ANOVA using `rstatix::eta_squared()`  

#### Review

- read in data using `read_csv()`  
- chain functions together using ` |> `  
- modify columns using `mutate()`  
- select columns using `select()`  
- rename columns using `rename()`  
- visualize data using `ggplot()`  
- create histograms using `geom_histogram()`  
- visualize QQ plots using `geom_qq()` and `geom_qq_line()`  
- create multi-panel plots using `facet_wrap()`  
- group data using `group_by()`  
- summarize data using `summarize()`  

### Data sources

The Plum Island Ecosystem fiddler crab data is from `lterdatasampler` (data info [here](https://lter.github.io/lterdatasampler/articles/pie_crab_vignette.html)). Crab carapace width is measured in mm.

## 2. Code

### 1. Packages

```{r read-packages}
#| eval: false

library(tidyverse)
library(lterdatasampler)
library(rstatix)
library(car)

# for parametric tests
data(pie_crab)
```

```{r data-for-website}
#| echo: false
#| message: false

library(tidyverse)
library(lterdatasampler)
library(rstatix)
library(car)

# dataset from lterdatasampler package
data(pie_crab)
```


### 2. Parametric tests

#### a. Cleaning and wrangling

- Create a new object called `pie_crab_clean`.
- Filter to only include the following sites: Cape Cod, Virginia Coastal Reserve LTER, and Zeke's Island NERR.
- Make sure `site` is a factor.
- Select the columns of interest: `site`, `name`, and `code`.
- Rename the column `name` to `site_name` and `code` to `site_code`. 

```{r crab-cleaning}
pie_crab_clean <- pie_crab |> # start with the pie_crab dataset
  filter(site %in% c("CC", "ZI", "VCR")) |> # filter for Cape Cod, Zeke's Island, Virginia Coastal
  mutate(name = as_factor(name)) |> # make sure site name is being read in as a factor
  select(site, name, size) |> # select columns of interest
  rename(site_code = site, # rename site to be site_code
         site_name = name) # rename name to be site_name

# display some rows from the data frame
# useful for showing a small part of the data frame
slice_sample(pie_crab_clean, # data frame
             n = 10) # number of rows to show
```

#### b. Quick summary

- Create a new object called `pie_crab_summary`.
- Calculate the mean, variance, and sample size. Display the object.

```{r crab-summary}
# creating a new object called pie_crab_summary
pie_crab_summary <- pie_crab_clean |> # starting with clean data frame
  group_by(site_name) |> # group by site
  summarize(mean = mean(size), # calculate mean size
            var = var(size), # calculate variance of size
            n = length(size)) |> # calculate number of observations per site (sample size)
  ungroup() # ungrouping

# display pie_crab_summary
pie_crab_summary
```

#### c. Exploring the data

Create a jitter plot with the mean crab carapace width for each site.

```{r crab-explore-viz}
# base layer: ggplot
ggplot(pie_crab_clean, # use the clean data set
       aes(x = site_name, # x-axis
           y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "point", # geometry being plotted
               fun = mean, # function (calculating the mean)
               color = "red", # color to make it easier to see
               size = 5) + # larger point size
  theme_minimal() # cleaner plot theme
```

Is there a difference in mean crab carapace width between the three sites?  

**Yes, and Zeke's Island NERR crabs tend to be smaller than those from Cape Cod or Virginia Coastal Reserve LTER.**

#### d. Check 1: normally distributed variable

- Create a histogram of crab size.
- Facet your histogram so that you have 3 panels, with one panel for each site.

```{r crab-histogram}
# base layer: ggplot
ggplot(data = pie_crab_clean, 
       aes(x = size)) + # x-axis
  # first layer: histogram
  geom_histogram(bins = 6, # bins from Rice Rule (approximate)
                 fill = "orange",
                 color = "black") + 
  # faceting by site_name: creating 3 different panels
  facet_wrap(~ site_name) 
```

- Create a QQ plot of crab size. 
- Facet your QQ plot so that you have 3 panels, with one panel for each site.

```{r crab-qq}
# base layer: ggplot
ggplot(data = pie_crab_clean, 
       aes(sample = size)) + # y-axis
  # first layer: QQ plot reference line
  geom_qq_line() + 
  # second layer: QQ plot points (actual observations)
  geom_qq(color = "orange") + 
  # faceting by site_name
  facet_wrap(~ site_name, 
             scales = "free") # let axes vary between panels
```

What are the outcomes of your visual checks?  

**Not perfect (crab carapace width from VCR LTER seems not normally distributed) but with large sample sizes for each group, this might not matter too much.**  

#### e. Check 2: equal variances

Do a gut check: is the ratio of the largest variance to the smallest variance less than 3?

```{r crab-variances}
8.63/4.04
```

What are the outcomes of your variance check?  

**By the rule of thumb (ratio of largest variance to smallest variance < 3), the variances are close enough; the largest variance is about 2.1 times larger than the smallest variance.**

#### f. ANOVA

```{r crab-anova}
# creating an object called crab_anova
crab_anova <- aov(size ~ site_name, # formula
                  data = pie_crab_clean) # data

# test object
crab_anova

# gives more information
summary(crab_anova)
```

Summarize results: is there a difference in mean crab carapace width between the three sites?  

**There is a difference in mean crab carapace width between Cape Cod, Virginia Coastal Reserve LTER, and Zeke's Island NERR (one-way ANOVA, F(2, 89) = 39.6, p < 0.001, $\alpha$ = 0.05).**

#### g. Post-hoc: Tukey HSD

```{r crab-tukey}
TukeyHSD(crab_anova)
```

Which pairwise comparisons are actually different? Which ones are not different?  

**Zeke's Island NERR and Cape Cod crabs are different, and Zeke's Island NERR and Virginia Coastal Reserve LTER crabs are different. Virginial Coastal Reserve LTER and Cape Cod crabs are not different.**

#### h. effect size

Using `eta_squared()` from `rstatix`

```{r crab-effectsize}
eta_squared(crab_anova)
```

What is the magnitude of the differences between sites in crab size?  

**There is a large difference in crab size between sites.**

#### i. Putting everything together

We found a large ($\eta^2$ = 0.47) difference between sites in mean crab carapace width (one-way ANOVA, F(2, 89) = 39.6, p < 0.001, $\alpha$ = 0.05). The smallest crabs were from Zeke's Island NERR, which were on average 12.1 mm. Zeke's Island crabs were 4.8 mm (95% CI: [3.3, 6.2] mm) smaller than crabs from Cape Cod (Tukey's HSD: p < 0.001) and 4.3 mm (95% CI: [2.9, 5.7] mm) smaller than crabs from Virginia Coastal Reserve LTER (Tukey's HSD: p < 0.001).  

### 3. Figures and captions

See your workshop materials for a PDF on how to write captions to accompany figures.

