---
title: "Coding workshop: Week 5"
subtitle: "Analysis of variance"
date: 2026-05-01
categories: [tidyverse, lterdatasampler, rstatix, aov, summary, TukeyHSD, eta_squared, read_csv, pipe operators, '|>', select, rename, ggplot, geom_histogram, geom_qq, geom_qq_line, facet_wrap, group_by, summarize, as_factor, mutate]
format:
  html:
    toc: true
    toc-depth: 8
---

[Workshop date: Friday 1 May]{style="color: #79ACBD; font-size: 24px;"}

## 1. Summary

### Packages

- `tidyverse`
- `lterdatasampler`
- `rstatix`

### Operations

#### New functions

- make sure a variable is a factor using `as_factor()`
- do an ANOVA using `aov()`
- look for more information from model results using `summary()`
- do post-hoc Tukey test using `TukeyHSD()`
- calculate effect size for ANOVA using `rstatix::eta_squared()`

#### Review

- read in data using `read_csv()`
- chain functions together using ` |> `
- modify columns using `mutate()`
- select columns using `select()`
- rename columns using `rename()`
- visualize data using `ggplot()`
- create histograms using `geom_histogram()`
- visualize QQ plots using `geom_qq()` and `geom_qq_line()`
- create multi-panel plots using `facet_wrap()`
- group data using `group_by()`
- undo grouping using `ungroup()`
- summarize data using `summarize()`

### Data sources

The Plum Island Ecosystem fiddler crab data is from `lterdatasampler` (data info [here](https://lter.github.io/lterdatasampler/articles/pie_crab_vignette.html)). Crab carapace width is measured in mm.

## 2. Code

### 1. Set up

#### Packages

```{r}
#| label: packages
#| message: false

library(tidyverse)
library(lterdatasampler)
library(effectsize)
library(rstatix)
```

#### Data 
```{r}
#| label: data-for-website
#| eval: false

# for parametric tests
data(pie_crab)

# for non-parametric tests
ramen_ratings <- read_csv("ramen_ratings.csv")
```

```{r}
#| label: data
#| echo: false
#| message: false

# for parametric tests
data(pie_crab)

# for non-parametric tests
ramen_ratings <- read_csv(here::here("workshop", "data", "ramen_ratings.csv"))
```

### 2. Parametric tests

#### a. Cleaning and wrangling

Create a new object called `pie_crab_clean`. Filter to only include the following sites: Cape Cod, Virginial Coastal Reserve LTER, and Zeke's Island NERR. Select the columns of interest: `site`, `name`, and `code`. Rename the column `name` to `site_name` and `code` to `site_code`. 

```{r}
#| label: crab-cleaning

# start with the pie_crab dataset
pie_crab_clean <- pie_crab |> 
  # filter for Cape Cod, Zeke's Island, Virginia Coastal
  filter(site %in% c("CC", "ZI", "VCR")) |> 
  # select columns of interest
  select(site, name, size) |> 
  # rename columns
  rename(site_code = site, 
         site_name = name) 
```

#### b. Quick summary

Create a new object called `pie_crab_summary`. Calculate the mean, variance, and sample size. Display the object.

```{r}
#| label: crab-summary

# creating a new object called pie_crab_summary
pie_crab_summary <- pie_crab_clean |> 
  # group by site
  group_by(site_name) |> 
  # calculate mean, variance, sample size
  summarize(mean = mean(size), 
            var = var(size), 
            n = length(size)) |> 
  # ungroup data frame
  ungroup()

# display pie_crab_summary
pie_crab_summary
```

#### c. Exploring the data

Create a jitter plot with the mean crab size for each site.

```{r}
#| label: exploratory visualization

# base layer: ggplot
ggplot(data = pie_crab_clean, 
       # x-axis: full site name
       # y-axis: crab size
       mapping = aes(x = site_name, 
                     y = size)) + 
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.1) + 
  # second layer: point representing mean
  stat_summary(geom = "point", # geometry being plotted
               fun = mean, # function (calculating the mean)
               color = "red", # color to make it easier to see
               size = 5) + # larger point size
  # cleaner plot theme
  theme_minimal() 
```

Is there a difference in mean crab carapace width (mm) between the three sites? What visual components of the plot are you using to make that guess? 

**insert response here**

#### d. Check 1: normally distributed variable

Create a histogram of crab size. Facet your histogram so that you have 3 panels, with one panel for each site.

```{r}
#| label: crab-histogram

# base layer: ggplot
ggplot(data = pie_crab_clean, 
       aes(x = size)) + # x-axis
  # first layer: histogram
  geom_histogram(bins = 9) + # number of bins from Rice Rule
  # faceting by site_name: creating 3 different panels
  facet_wrap(~ site_name) 
```

Create a QQ plot of crab size. Facet your QQ plot so that you have 3 panels, with one panel for each site.

```{r}
#| label: crab-qq

# base layer: ggplot
ggplot(data = pie_crab_clean, 
       aes(sample = size)) + # y-axis
  # first layer: QQ plot reference line
  geom_qq_line(color = "orange") + 
  # second layer: QQ plot points (actual observations)
  geom_qq() + 
  # faceting by site_name
  facet_wrap(~ site_name, 
             scales = "free") # let axes vary between panels
```

What are the outcomes of your visual checks?  

**summarize outcomes here**  

#### e. Check 2: equal variances

Do a gut check: is the ratio of the largest variance to the smallest variance less than 3?

```{r}
#| label: crab-variance

8.63/4.04
```

#### f. ANOVA

```{r}
#| label: crab-anova

# creating an object called crab_anova
crab_anova <- aov(
  # size as a function of site
  size ~ site_name, 
  # using clean data frame
  data = pie_crab_clean
) 

# object
crab_anova

# gives more information
summary(crab_anova)
```

Summarize results: is there a difference in mean crab size (measured as carapace width in mm) between the three sites?  

**There is a difference in mean crab carapace width between Cape Cod, Virginia Coastal Reserve LTER, and Zeke's Island NERR (one-way ANOVA, F(2, 89) = 39.6, p < 0.001, $\alpha$ = 0.05).**

#### g. Post-hoc: Tukey HSD

```{r}
#| label: crab-tukey

TukeyHSD(crab_anova)
```

Which pairwise comparisons are actually different? Which ones are not different?  

**Zeke's Island NERR and Cape Cod crabs are different, and Zeke's Island NERR and Virginia Coastal Reserve LTER crabs are different. Virginia Coastal Reserve LTER and Cape Cod crabs are not different.**

#### h. effect size

Using `eta_squared()` from `rstatix`

```{r}
#| label: crab-effectsize

rstatix::eta_squared(crab_anova)
```

What is the magnitude of the differences between sites in crab size?  

**There is a large difference in crab size between sites.**  

Another way of saying this: there is a **large** effect of site on crab size. There is a **large** difference between sites in crab size.

#### i. Putting everything together

**means**

Zeke's Island NERR: 12.1 mm
Cape Cod: 16.8 mm
Virginia Coastal Reserve LTER: 16.3 mm

**test type**

one-way ANOVA

**test distribution**

F

**distribution parameters**

numerator and denominator degrees of freedom: 2, 89

**test statistic**

F-statistic: 39.6

**p-value**

5.1e-13, expressed as p < 0.001

**significance level**

$\alpha$ = 0.05

**post-hoc tests**

used Tukey HSD (Honestly Significant Difference)
Zeke's Island different from Cape Cod and VCR, Cape Cod and VCR are not different

**effect size**

$\eta^2$ = 0.47

**Writing:**

We found a large ($\eta^2$ = 0.47) difference between sites in mean crab carapace width (one-way ANOVA, F(2, 89) = 39.6, p < 0.001, $\alpha$ = 0.05). The smallest crabs were from Zeke's Island NERR, which had an average width of 12.1 mm. Zeke's Island crabs were 4.8 mm (95% CI: [3.3, 6.2] mm) smaller than crabs from Cape Cod (Tukey's HSD: p < 0.001) and 4.3 mm (95% CI: [2.9, 5.7] mm) smaller than crabs from Virginia Coastal Reserve LTER (Tukey's HSD: p < 0.001).  

### 3. Non-parametric tests

#### a. Clean and wrangle the data

```{r}
#| label: ramen-wrangle

# create new object from ramen ratings
ramen_ratings_clean <- ramen_ratings |> 
  # filter to only include Maruchan ramen
  filter(brand == "Maruchan") |> 
  # reorder style factor
  mutate(style = fct_relevel(style, "Bowl", "Pack", "Tray", "Cup")) 

# look at the structure
str(ramen_ratings_clean)
```

#### b. Quick summary

```{r}
#| label: ramen-summary

# create a new object from clean data frame
ramen_ratings_summary <- ramen_ratings_clean |> 
  # group by style
  group_by(style) |> 
  # calculate the median
  summarize(median = median(stars)) |> 
  # ungroup data frame
  ungroup()

# display the object
ramen_ratings_summary 
```


#### c. Make a boxplot to compare star ratings across ramen styles

```{r}
#| label: ramen-exploratory-visualization

# base layer: ggplot
ggplot(data = ramen_ratings_clean, 
       # x-axis: ramen style
       # y-axis: star rating
       # color by style
       mapping = aes(x = style, 
                     y = stars, 
                     color = style)) + 
  # first layer: boxplot
  geom_boxplot(color = "darkgrey", 
               # removing outliers because they're shown in jitter
               outliers = FALSE) + 
  # second layer: jitter to represent individual ramen styles
  geom_jitter(height = 0,
              width = 0.1,
              size = 2,
              alpha = 0.6) +
  # set custom colors
  scale_color_manual(values = c("firebrick4", "orange", "gold", "darkgreen")) + 
  # minimal theme
  theme_minimal() + 
  # take out the legend
  theme(legend.position = "none") 
```

#### d. Do the Kruskal-Wallis test

```{r}
#| label: ramen-kw

kruskal.test(
  # formula: star ratings as a function of ramen style
  stars ~ style, 
  # data frame
  data = ramen_ratings_clean 
)
```

Is there a difference in ratings between ramen styles?  

**Ramen styles differ in ratings (Kruskal-Wallis rank sum test, $\chi^2$(3) = 15.7, p = 0.0013).**

#### e. Do a Dunn's post-hoc test

Using `dunn_test()` from `rstatix`

```{r}
#| label: ramen-post-hoc

dunn_test(
  # formula: star ratings as a function of ramen style
  stars ~ style, 
  # data frame
  data = ramen_ratings_clean 
)
```

Which pairwise comparisons of ramen styles are different from each other?  

**Bowls and packs are different, bowls and trays are different, bowls and cups are different. Packs are not different from trays or cups, and trays are not different from cups.**

#### f. Calculate an effect size

Using `kruskal_effsize()` from `rstatix`

```{r}
#| label: ramen-effectsize

kruskal_effsize(
  # formula: star ratings as a function of ramen style
  stars ~ style, 
  # data frame
  data = ramen_ratings_clean 
)
```

What is the magnitude of the effect of ramen style on ratings?  

**There is a moderate ($\eta^2$ = 0.12) effect of ramen style on ratings.**

#### g. Putting everything together

**medians**

Bowl: 4.25 stars
Pack: 3.75 stars
Tray: 3.75 stars
Cup: 3.5 stars

**test**

Kruskal-Wallis rank sum

**test "distribution"**

$\chi^2$

**distribution parameters**

degrees of freedom (3)

**test statistic**

15.7

**p-value**

0.0013

**significance level**

$\alpha$ = 0.05

**post-hoc test**

Dunn's post-hoc test with Holm p-value adjustment
Bowls are different from packs, trays, and cups

We found a moderate ($\eta^2$ = 0.12) difference in ratings between ramen styles (Kruskal-Wallis rank sum test, $\chi^2$(3) = 15.7, p = 0.0013, $\alpha$ = 0.05). Bowl-style ramen had a median rating of 4.25 stars, which tended to be more highly rated than pack (Dunn's post-hoc test: Holm adjusted p = 0.04, median rating = 3.75 stars), tray (Dunn's post-hoc test: Holm adjusted p = 0.04, median rating = 3.75 stars), or cup style ramen (Dunn's post-hoc test: Holm adjusted p = 0.0013, median rating = 3.5 stars).



### 4. Captions

#### Figure 1.

**Figure 1. Crab carapace width differs across sites.** Small black points represent measurements of crab carapace widths at three sites: Zeke's Island National Estuarine Research Reserve (NERR), Virginia Coastal Reserve Long-Term Ecological Research (LTER) site, and Cape Cod. Large red points represent mean crab carapace width. Data originally from: Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb. Accessed via `lterdatasampler` package.

#### Figure 2. 

**Figure 2. Histograms of crab carapace width.** Columns represent counts of individual crabs. Distributions of crab carapace width are represented in three panels: Zeke's Island NERR (left), Virginia Coastal Reserve LTER (center), and Cape Cod (right). Data originally from: Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb. Accessed via `lterdatasampler` package.  

Alternative option if you have a figure made using the same dataset as a previous figure:  

**Figure 2. Histograms of crab carapace widths.** Bars represent counts of crab sizes. Distributions are shown for three sites: Zeke's Island NERR (left), Virginia Coastal Reserve (center), and Cape Cod (right). Data source and access as in Figure 1. 

#### Figure 3.

**Figure 3. Quantile-Quantile (QQ) plots of crab carapace width.** Points represent individual observations of crab carapace width. Lines represent reference line relating theoretical quantiles (x-axis) and sample quantiles (y-axis). Panels represent different sites: Zeke's Island NERR (left), Virginia Coastal Reserve LTER (center), and Cape Cod (right). Data originally from: Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb. Accessed via `lterdatasampler` package.  

Alternative option if you have a figure made using the same dataset as a previous figure: 

**Figure 3. Quantile-quantile (QQ) plots of crab carapace widths.** Points represent individual crab carapace width measurements. Lines represent reference quantiles of sample (y-axis) and standard normal theoretical (x-axis) distributions. Panels, data source, and access as in Figure 2.

#### Figure 4. 

**Figure 4. Median star ratings for Maruchan Ramen styles.** Points represent individual ratings for ramen types in each style (red: bowl, orange: pack, yellow: tray, green: cup). Boxplots represent median star ratings in each style.

**END OF WORKSHOP 5**

## Extra stuff

### More applications of `stat_summary()`

`stat_summary()` in the `ggplot2` package is a useful function to do quick summary visualizations within a plot.  

In class, we used `stat_summary()` to plot mean carapace widths at each site in red (see the "Exploring the data" section above).  

You can also use `stat_summary()` to visualize means and bars (representing standard deviation, standard error, confidence intervals, etc.).  

For example, here is a plot using `stat_summary()` to visualize means and standard errors:

```{r}
#| label: mean and SD

# base layer: ggplot
ggplot(data = pie_crab_clean, # use the clean data set
       mapping = aes(x = site_name, # x-axis
                     y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "pointrange", # geometry being plotted
               fun.data = mean_sdl, # function to calculate mean and sd (note new argument name)
               color = "red", # color to make it easier to see
               size = 1) + # larger point size
  theme_minimal() + # cleaner plot theme
  labs(x = "Site name",
       y = "Carapace width (mm)",
       caption = "Mean \U00B1 standard deviation")

```

And standard error:

```{r}
#| label: mean and SE

# base layer: ggplot
ggplot(data = pie_crab_clean, # use the clean data set
       mapping = aes(x = site_name, # x-axis
                     y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "pointrange", # geometry being plotted
               fun.data = mean_se, # function to calculate mean and SE (note new argument name)
               color = "red", # color to make it easier to see
               size = 1) + # larger point size
  theme_minimal() + # cleaner plot theme
  labs(x = "Site name",
       y = "Carapace width (mm)",
       caption = "Mean \U00B1 standard error")

```

And a 95% confidence interval:

```{r}
#| label: mean and confidence interval

# base layer: ggplot
ggplot(data = pie_crab_clean, # use the clean data set
       mapping = aes(x = site_name, # x-axis
                     y = size)) + # y-axis
  # first layer: jitter (each point is an individual crab)
  geom_jitter(height = 0, # don't jitter points vertically
              width = 0.15) + # narrower jitter (easier to see)
  # second layer: point representing mean
  stat_summary(geom = "pointrange", # geometry being plotted
               fun.data = mean_cl_normal, # function to calculate range for confidence interval (note new argument name)
               color = "red", # color to make it easier to see
               size = 1) + # larger point size
  theme_minimal() + # cleaner plot theme
  labs(x = "Site name",
       y = "Carapace width (mm)",
       caption = "Means with 95% confidence interval")

```









